### 2024-09-20 paper/paper.qmd
Saving the current results and plots to analyse later.

---
title: "Analysis of Crime Statistics from 2014 to 2023 in Neighborhoods of Varying Home Price"
subtitle: "A Statistical Basis for Policy Improvement"
author: Chenika Bukes
thanks: "Code and data are available at: https://github.com/chenikabukes/TorontoDataset"
date: 24/09/2024
date-format: long
abstract: "The abstract answers: 1) what was done, 2) what was found, and 3) why this matters (all at a high level). Likely four sentences. Abstract must make clear what we learn about the world because of this paper."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library("tidyverse")
library("gt")

```

# Introduction

Housing prices and crime rates are often perceived as indicators of the socio-economic environment of a city’s neighborhoods. Higher home prices are typically associated with safer, more affluent areas, while lower home prices may correlate with higher crime rates and other social challenges. However, these relationships are not always straightforward and may vary across different regions and time periods.

This report aims to investigate the relationship between housing prices and crime rates in various neighborhoods of Toronto from 2014 to 2023. By focusing on the 10 neighborhoods with the highest home prices and the 10 neighborhoods with the lowest home prices, this analysis seeks to identify any significant patterns or correlations between housing prices and crime rates over time.

Through statistical analysis and visualizations, we aim to answer the following key questions:

How have crime rates changed in Toronto’s neighborhoods over the last decade?
Are neighborhoods with higher home prices experiencing lower crime rates, or is the relationship more complex?
What types of crimes are most prevalent in high-price vs. low-price neighborhoods?
The findings from this analysis will provide insights into the socio-economic dynamics of Toronto's neighborhoods and serve as a basis for city planners and policymakers in addressing crime prevention strategies.

# Data {#sec-data}
# Raw Data {#sub-section}
The data for this analysis comes from two key sources within Toronto's Open Data portal:

**Housing Price Data**: This dataset contains home prices for various neighborhoods in Toronto. The data includes details such as neighborhood name, neighborhood ID, and the average home price for each area. The housing data is extracted from the “Wellbeing Toronto” dataset, which provides information on home prices in Toronto’s neighborhoods. The raw dataset contains multiple columns representing various housing and socio-economic indicators. For the purpose of this analysis, the column of interest is the average home price for each neighborhood, which has been cleaned and filtered for further analysis.

**Crime Statistics Data**: This dataset provides crime statistics for each neighborhood, with detailed rates for various types of crimes, including:
  - Assault
  - Auto theft
  - Bike theft
  - Break and enter
  - Homicide
  - Robbery
  - Shooting
  - Theft from motor vehicle
  - Theft over $5,000
The crime data spans from 2014 to 2023 and includes crime rates for each of these categories on an annual basis. The crime rates are reported per 100,000 residents, allowing for comparisons across neighborhoods with different population sizes.

**Data Cleaning and Preparation**
To prepare the data for analysis, the following steps were taken:

  - Housing Data: The home price data was cleaned by removing any invalid or missing entries, converting home prices to numeric values, and filtering out zero or negative prices. The final dataset consists of neighborhoods and their corresponding home prices.

  - Crime Data: The crime data was cleaned by filtering out any missing or anomalous entries. The crime rates for each neighborhood were then aggregated by year and by crime type to allow for time-series analysis and comparisons across neighborhoods.

**Focus Neighborhoods**
The neighborhoods with the 10 highest and 10 lowest home prices were selected for focused analysis. These neighborhoods represent a broad spectrum of socio-economic conditions and are expected to exhibit varying crime patterns.

Explain if there were similar datasets that could have been used and why they were not. If variables were constructed then this should be mentioned, and high-level cleaning aspects of note should be mentioned, but this section should focus on the destination, not the journey. It is important to understand what the variables look like by including graphs, and possibly tables, of all observations, along with discussion of those graphs and the other features of these data. Summary statistics should also be included, and well as any relationships between the variables. If this becomes too detailed, then appendices could be used.

Below is an overview of the housing dataset, presented in Table 1, and visualizations that illustrate key trends in the data.

```{r}
#| label: tbl-housing-summary
#| tbl-cap: "Table 1: Summary Statistics of Housing Price Data (Top 10 Highest and Lowest)"
#| echo: false

# Read in the housing data
housing_cleaned_data <- read_csv("../data/clean_data/housing_cleaned_data.csv")

# Summary for the top 10 highest and lowest home prices
housing_summary <- housing_cleaned_data %>%
  summarise(
    Mean_Price = mean(price),
    Median_Price = median(price),
    Max_Price = max(price),
    Min_Price = min(price)
  )

# Display the summary table using gt
housing_summary_gt <- housing_summary %>%
  gt() %>%
  tab_header(
    title = md("**Summary Statistics of Housing Price Data (Top 10 Highest and Lowest)**")
  )

housing_summary_gt

```
```{r}
#| label: tbl-crime-summary
#| tbl-cap: "Table 2: Summary Statistics of Crime Rates (2014-2023)"
#| echo: false

# Read in the crime data
crime_cleaned_data <- read_csv("../data/clean_data/crime_cleaned_data.csv")

# Summarise crime data by calculating the mean for each crime rate across all years
crime_summary <- crime_cleaned_data %>%
  summarise(
    Mean_Assault_Rate = rowMeans(select(., starts_with("assault_rate")), na.rm = TRUE),
    Mean_Autotheft_Rate = rowMeans(select(., starts_with("autotheft_rate")), na.rm = TRUE),
    Mean_Biketheft_Rate = rowMeans(select(., starts_with("biketheft_rate")), na.rm = TRUE),
    Mean_Breakenter_Rate = rowMeans(select(., starts_with("breakenter_rate")), na.rm = TRUE),
    Mean_Homicide_Rate = rowMeans(select(., starts_with("homicide_rate")), na.rm = TRUE),
    Mean_Robbery_Rate = rowMeans(select(., starts_with("robbery_rate")), na.rm = TRUE),
    Mean_Shooting_Rate = rowMeans(select(., starts_with("shooting_rate")), na.rm = TRUE),
    Mean_Theftfrommv_Rate = rowMeans(select(., starts_with("theftfrommv_rate")), na.rm = TRUE),
    Mean_Theftover_Rate = rowMeans(select(., starts_with("theftover_rate")), na.rm = TRUE)
  )

# Display the summary table using gt
crime_summary_gt <- crime_summary %>%
  gt() %>%
  tab_header(
    title = md("**Summary Statistics of Crime Rates (2014-2023)**")
  )

crime_summary_gt

```

**Discussion of Data Selection**
I chose this dataset over the census data because...
# Results {#sec-results}
# Crime Rate Changes over Time {#subsection}
Crime rate trends over the years (2014-2023) were analyzed for each crime type to understand patterns and shifts. This is visualized in the following plots:
```{r}
#| label: fig-transport-vs-stationary
#| fig-cap: "Figure 1: Time-series visualization of Crime Rates from 2014-2023"
#| echo: false

crime_cleaned_data_long <- crime_cleaned_data %>%
  pivot_longer(
    cols = contains("rate"),  
    names_to = "Crime_Type",   
    values_to = "Rate"         
  ) %>%
  separate(Crime_Type, into = c("Crime", "Year"), sep = "_rate_")

# Plot the time-series visualization of crime rates from 2014-2023
crime_cleaned_data_long %>%
  ggplot(aes(x = Year, y = Rate, color = Crime)) +
  geom_line() +
  facet_wrap(~ Crime, scales = "free_y") +
  labs(title = "Crime Rate Trends in Toronto (2014-2023)",
       x = "Year",
       y = "Crime Rate per 100,000 Residents") +
  theme_minimal()

```

```{r}
#| label: fig-percentile-crime-rates
#| fig-cap: "Figure 2: Box Plots comparing Crime Rates in Top 10 vs Bottom 10 House Price Neighbourhoods"
#| echo: false

# Check column names of housing_cleaned_data
names(housing_cleaned_data)

# Check column names of crime_cleaned_data
names(crime_cleaned_data)

# Select the top 10 highest and lowest priced neighborhoods
top_10_highest <- housing_cleaned_data %>%
  top_n(10, price)

top_10_lowest <- housing_cleaned_data %>%
  top_n(-10, price)

# Add 'neighbourhood_type' to label the neighbourhoods
top_10_highest <- top_10_highest %>%
  mutate(neighbourhood_type = "Top 10 Highest")

top_10_lowest <- top_10_lowest %>%
  mutate(neighbourhood_type = "Bottom 10 Lowest")

# Combine the two datasets (top and bottom neighborhoods)
top_bottom_neighborhoods <- bind_rows(top_10_highest, top_10_lowest)

# Filter the crime data for those top 10 highest and lowest priced neighborhoods
crime_high_low <- crime_cleaned_data %>%
  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code)  # Make sure 'neighbourhood_code' is used here

# Join the filtered crime data with the top/bottom housing data
# Create an overall crime rate by summing all crime rate columns
crime_high_low <- crime_cleaned_data %>%
  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code) %>%
  mutate(overall_crime_rate = rowSums(select(., contains("rate")), na.rm = TRUE))  # Sum of all crime rates

# Join with housing data to get neighborhood type labels
crime_high_low <- crime_high_low %>%
  left_join(top_bottom_neighborhoods, by = c("hood_id" = "neighbourhood_code"))

# Box plot comparing overall crime rates between high and low-priced neighborhoods
ggplot(crime_high_low, aes(x = neighbourhood_type, y = overall_crime_rate, fill = neighbourhood_type)) +
  geom_boxplot() +
  labs(title = "Overall Crime Rate Comparison Between High and Low-Priced Neighborhoods",
       x = "Neighborhood Type",
       y = "Overall Crime Rate per 100,000 Residents") +
  theme_minimal()

```

Box Plots Comparing Crime Rates for High vs. Low Price Neighborhoods:

This helps visualize how crime rates differ between the top 10 highest and lowest-priced neighborhoods.

```{r}

```

Bar Plots Comparing Crime Rates Between High and Low Price Neighborhoods:
This will provide a clear comparison between the two groups for each type of crime.

```{r}
#| label: fig-transport-vs-stationary
#| fig-cap: "Figure 2: Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
#| echo: false

# Bar plot comparing crime rates for top 10 highest vs. lowest price neighborhoods
# First, pivot the crime data longer to create a year column
crime_high_low_long <- crime_cleaned_data %>%
  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code) %>%
  pivot_longer(cols = contains("rate"), 
               names_to = "Crime_Type_Year", 
               values_to = "Rate")

# Separate the Crime_Type_Year column into 'Crime_Type' and 'Year'
crime_high_low_long <- crime_high_low_long %>%
  separate(Crime_Type_Year, into = c("Crime_Type", "Year"), sep = "_rate_")

# Convert the 'Year' column to numeric
crime_high_low_long <- crime_high_low_long %>%
  mutate(Year = as.numeric(Year))

# Calculate the overall crime rate for each neighborhood and year
# Using either rowSums (for summing) or rowMeans (for averaging)
crime_high_low_yearly <- crime_high_low_long %>%
  group_by(hood_id, Year) %>%
  summarise(overall_crime_rate = sum(Rate, na.rm = TRUE)) %>%
  ungroup()

# Join with housing data to get neighborhood type labels
crime_high_low_yearly <- crime_high_low_yearly %>%
  left_join(top_bottom_neighborhoods, by = c("hood_id" = "neighbourhood_code"))

# Group by neighborhood type and year, then calculate the average overall crime rate
crime_yearly_summary <- crime_high_low_yearly %>%
  group_by(neighbourhood_type, Year) %>%
  summarise(avg_overall_crime_rate = mean(overall_crime_rate, na.rm = TRUE)) %>%
  ungroup()

# Bar plot comparing overall crime rates between top and bottom-priced neighborhoods for each year
ggplot(crime_yearly_summary, aes(x = Year, y = avg_overall_crime_rate, fill = neighbourhood_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Overall Crime Rate for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
       x = "Year",
       y = "Average Overall Crime Rate per 100,000 Residents") +
  theme_minimal() +
  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))

```
```{r}
# First, pivot the crime data longer to create a year column
crime_high_low_long <- crime_cleaned_data %>%
  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code) %>%
  pivot_longer(cols = contains("rate"), 
               names_to = "Crime_Type_Year", 
               values_to = "Rate")

# Separate the Crime_Type_Year column into 'Crime_Type' and 'Year'
crime_high_low_long <- crime_high_low_long %>%
  separate(Crime_Type_Year, into = c("Crime_Type", "Year"), sep = "_rate_")

# Convert the 'Year' column to numeric
crime_high_low_long <- crime_high_low_long %>%
  mutate(Year = as.numeric(Year))

# Join with housing data to get neighborhood type labels
crime_high_low_long <- crime_high_low_long %>%
  left_join(top_bottom_neighborhoods, by = c("hood_id" = "neighbourhood_code"))
### Bar Plot for Assault Rates ###

# Filter for assault rates
assault_rates <- crime_high_low_long %>%
  filter(Crime_Type == "assault") %>%
  group_by(neighbourhood_type, Year) %>%
  summarise(avg_assault_rate = mean(Rate, na.rm = TRUE)) %>%
  ungroup()

# Create bar plot for assault rates
ggplot(assault_rates, aes(x = Year, y = avg_assault_rate, fill = neighbourhood_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Assault Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
       x = "Year",
       y = "Average Assault Rate per 100,000 Residents") +
  theme_minimal() +
  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))
```

```{r}

### Bar Plot for Motor Vehicle Theft Rates (Autotheft) ###

# Filter for motor vehicle theft (autotheft) rates
autotheft_rates <- crime_high_low_long %>%
  filter(Crime_Type == "autotheft") %>%
  group_by(neighbourhood_type, Year) %>%
  summarise(avg_autotheft_rate = mean(Rate, na.rm = TRUE)) %>%
  ungroup()

# Create bar plot for motor vehicle theft rates
ggplot(autotheft_rates, aes(x = Year, y = avg_autotheft_rate, fill = neighbourhood_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Motor Vehicle Theft Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
       x = "Year",
       y = "Average Motor Vehicle Theft Rate per 100,000 Residents") +
  theme_minimal() +
  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))
```
```{r}
### Bar Plot for Shooting Rates ###

# Filter for shooting rates
assault_rates <- crime_high_low_long %>%
  filter(Crime_Type == "shooting") %>%
  group_by(neighbourhood_type, Year) %>%
  summarise(avg_assault_rate = mean(Rate, na.rm = TRUE)) %>%
  ungroup()

# Create bar plot for assault rates
ggplot(assault_rates, aes(x = Year, y = avg_assault_rate, fill = neighbourhood_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Shooting Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
       x = "Year",
       y = "Average Shooting Rate per 100,000 Residents") +
  theme_minimal() +
  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))

```
```{r}
### Bar Plot for Break and Enter Rates ###

# Filter for shooting rates
assault_rates <- crime_high_low_long %>%
  filter(Crime_Type == "breakenter") %>%
  group_by(neighbourhood_type, Year) %>%
  summarise(avg_assault_rate = mean(Rate, na.rm = TRUE)) %>%
  ungroup()

# Create bar plot for assault rates
ggplot(assault_rates, aes(x = Year, y = avg_assault_rate, fill = neighbourhood_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Break and Enter Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
       x = "Year",
       y = "Average Break and Enter Rate per 100,000 Residents") +
  theme_minimal() +
  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))

```
```{r}
### Bar Plot for Robbery Rates ###

# Filter for shooting rates
assault_rates <- crime_high_low_long %>%
  filter(Crime_Type == "robbery") %>%
  group_by(neighbourhood_type, Year) %>%
  summarise(avg_assault_rate = mean(Rate, na.rm = TRUE)) %>%
  ungroup()

# Create bar plot for assault rates
ggplot(assault_rates, aes(x = Year, y = avg_assault_rate, fill = neighbourhood_type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Robbery Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
       x = "Year",
       y = "Average Robbery Rate per 100,000 Residents") +
  theme_minimal() +
  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))

```
Figure 2 breaks down the number of clients transported to shelters versus those who remained stationary on buses at Spadina Station, providing a deeper understanding of how the initiative's resources were utilized.

The number of

# Model

The goal of our modeling strategy is twofold. Firstly, we aim to understand the factors driving the demand for transportation to warming centers versus the decision to remain stationary. Here, we briefly describe the Bayesian analysis model used to investigate these patterns. Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of clients using the service on day $i$. We model the relationship between $y_i$ and weather conditions using a Bayesian regression model:

\begin{align} y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \ \mu_i &= \alpha + \beta_1 \times \text{temperature}_i + \beta_2 \times \text{precipitation}_i \ \alpha &\sim \mbox{Normal}(0, 2.5) \ \beta_1, \beta_2 &\sim \mbox{Normal}(0, 2.5) \ \sigma &\sim \mbox{Exponential}(1) \end{align}

We run the model in R [@citeR] using the rstanarm package from @rstanarm. We use the default priors from rstanarm.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

### Model justification

We expect a positive relationship between adverse weather conditions (e.g., lower temperatures and higher precipitation) and the number of clients using the service. This is based on the assumption that more individuals will seek shelter during harsher weather conditions.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory Models of Client Usage Based on Weather Conditions"
#| warning: false

library(rstanarm)

first_model <- readRDS(file = here::here("models/first_model.rds"))

modelsummary::modelsummary(
  list(
    "First Model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)

```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-09-20 paper/paper.qmd
Updated paper to include analysis of the new crime and home price datasets.

@@
 thanks: "Code and data are available at: https://github.com/chenikabukes/TorontoDataset"
-date: 24/09/2024
+date: 24 September 2024
 date-format: long
-abstract: "The abstract answers: 1) what was done, 2) what was found, and 3) why this matters (all at a high level).
-Likely four sentences.
-Abstract must make clear what we learn about the world because of this paper."
-format: pdf
+format: 
+  pdf:
+    toc: true    
+    toc-depth: 3 
+    number-sections: true  
+    toc-title: "Table of Contents" 
+abstract: "This study analyzes crime rates across Toronto's neighborhoods from 2014 to 2023, focusing on the 10 neighborhoods with the highest home prices and the 10 neighborhoods with the lowest home prices.
+We investigated crime trends across nine categories, including assault, robbery, and autotheft, comparing the average crime rates between high-price and low-price neighborhoods.
+The analysis reveals a significant disparity in crime rates, with lower-priced neighborhoods consistently exhibiting higher crime rates, especially in violent crimes such as shootings and robberies.
+These findings underscore the importance of targeted crime prevention strategies in economically disadvantaged neighborhoods and suggest that housing affordability is closely linked to public safety concerns."
 number-sections: true
 bibliography: references.bib
+
 ---
@@
 #| include: false
+#| message: false
 #| warning: false
-#| message: false
+#| error: false
 
@@
 library("gt")
-
+library("scales")
 ```
@@
 
-Housing prices and crime rates are often perceived as indicators of the socio-economic environment of a city’s neighborhoods.
-Higher home prices are typically associated with safer, more affluent areas, while lower home prices may correlate with higher crime rates and other social challenges.
-However, these relationships are not always straightforward and may vary across different regions and time periods.
+We run the model in R [@citeR].
+Understanding the relationship between housing prices and crime rates is critical for urban planning and public policy.
+Housing prices are often considered a key indicator of socio-economic status, with higher-priced neighborhoods typically associated with lower crime rates, greater economic stability, and improved social services.
+Conversely, lower-priced neighborhoods are often linked to higher crime rates, economic challenges, and reduced access to resources.
+However, while these patterns may hold in many cases, the relationship between housing prices and crime rates is complex and not universally consistent across all regions.
+This complexity raises the need for more detailed investigation into how these factors interact in specific urban settings, such as Toronto.
 
-This report aims to investigate the relationship between housing prices and crime rates in various neighborhoods of Toronto from 2014 to 2023.
-By focusing on the 10 neighborhoods with the highest home prices and the 10 neighborhoods with the lowest home prices, this analysis seeks to identify any significant patterns or correlations between housing prices and crime rates over time.
+Despite numerous studies exploring socio-economic indicators and crime, there remains a gap in understanding how housing prices, particularly in high- and low-income neighborhoods, correlate with different types of crimes over time.
+Previous research often treats these neighborhoods as static entities, overlooking the dynamic nature of crime trends and their fluctuations over time.
+This paper aims to fill this gap by analyzing the crime rates across Toronto’s neighborhoods from 2014 to 2023, specifically focusing on the 10 neighborhoods with the highest and the 10 neighborhoods with the lowest home prices.
 
-Through statistical analysis and visualizations, we aim to answer the following key questions:
+In this analysis, we examined nine distinct crime categories—including assault, robbery, and autotheft—and compared the trends in both high- and low-priced neighborhoods.
+The goal was to identify significant patterns in crime rates, understand whether wealthier neighborhoods consistently experience lower crime rates, and investigate which types of crimes are most prevalent in different socio-economic contexts.
+Our findings reveal notable disparities, particularly in violent crimes such as shootings and robberies, which were found to be more prevalent in lower-income neighborhoods.
 
-How have crime rates changed in Toronto’s neighborhoods over the last decade?
-Are neighborhoods with higher home prices experiencing lower crime rates, or is the relationship more complex?
-What types of crimes are most prevalent in high-price vs. low-price neighborhoods?
-The findings from this analysis will provide insights into the socio-economic dynamics of Toronto's neighborhoods and serve as a basis for city planners and policymakers in addressing crime prevention strategies.
+These findings are important because they offer insights that can help city planners and policymakers design targeted crime prevention strategies.
+By understanding how crime rates differ across neighborhoods with varying economic conditions, more effective resource allocation and policy interventions can be implemented.
+The structure of this paper is as follows: first, we review the existing literature on the relationship between housing prices and crime.
+Second, we present our methodology and data sources.
+Third, we provide a detailed analysis of the results, followed by a discussion of the implications for public policy and future research.
 
 # Data {#sec-data}
-# Raw Data {#sub-section}
-The data for this analysis comes from two key sources within Toronto's Open Data portal:
+##  Raw Data {#sub-section}
+The data for this analysis comes from two primary sources obtained from Toronto’s Open Data portal:
 
-**Housing Price Data**: This dataset contains home prices for various neighborhoods in Toronto.
-The data includes details such as neighborhood name, neighborhood ID, and the average home price for each area.
-The housing data is extracted from the “Wellbeing Toronto” dataset, which provides information on home prices in Toronto’s neighborhoods.
-The raw dataset contains multiple columns representing various housing and socio-economic indicators.
-For the purpose of this analysis, the column of interest is the average home price for each neighborhood, which has been cleaned and filtered for further analysis.
+**Housing Price Data**: This dataset includes information on average home prices across various neighborhoods in Toronto.
+It serves as an indicator of the socio-economic status of each neighborhood.
+The data is extracted from the "Wellbeing Toronto" dataset, which records various socio-economic variables across the city.
+For this analysis, the focus is on the average home price for each neighborhood, identified by the neighborhood name and ID.
+This variable acts as a proxy for socio-economic conditions, with higher home prices suggesting wealthier neighborhoods, and lower prices implying economically disadvantaged areas.
 
-**Crime Statistics Data**: This dataset provides crime statistics for each neighborhood, with detailed rates for various types of crimes, including:
+**Crime Statistics Data**: This dataset contains detailed annual crime statistics for each neighborhood between 2014 and 2023.
+It includes crime rates per 100,000 residents for a range of criminal activities:
   - Assault
@@
   - Theft over $5,000
-The crime data spans from 2014 to 2023 and includes crime rates for each of these categories on an annual basis.
-The crime rates are reported per 100,000 residents, allowing for comparisons across neighborhoods with different population sizes.
+The crime data provides a robust view of criminal activity across various categories, offering insight into both violent and non-violent crime trends in different neighborhoods.
 
-**Data Cleaning and Preparation**
-To prepare the data for analysis, the following steps were taken:
+**Data Selection Reasoning**
+While Toronto’s census data also contains valuable socio-economic information, the housing price dataset was chosen because it was updated more frequently and provides clearer spatial links between neighborhoods and socio-economic status.
+The census data, recorded in 2015 and 2021, presented two key limitations for this analysis: (1) It was not yet fully updated to 2021 data, and (2) the census dataset organizes data by individual household income rather than neighborhood-level data, making it more challenging to compare neighborhoods directly.
+The housing price data, already grouped by neighborhood code, was easier to integrate with the crime statistics for this study's purpose.
 
-  - Housing Data: The home price data was cleaned by removing any invalid or missing entries, converting home prices to numeric values, and filtering out zero or negative prices.
-The final dataset consists of neighborhoods and their corresponding home prices.
+**Variables and their Roles**
+The key variables under examination are:
+  - Neighborhood Name and ID: Each neighborhood is identified by both a name and a unique ID code.
+  - Home Prices: Average home prices (in CAD) across neighborhoods, reflecting the relative affluence of each area.
+  - Crime Rates: Crime statistics for nine different crime categories (listed above), measured as the number of incidents per 100,000 residents annually.
+These crime rates provide a normalized metric for comparing areas with different population sizes.
+These variables allow for a detailed analysis of the relationship between socio-economic status (as proxied by home prices) and crime rates, both across crime categories and over time.
 
-  - Crime Data: The crime data was cleaned by filtering out any missing or anomalous entries.
-The crime rates for each neighborhood were then aggregated by year and by crime type to allow for time-series analysis and comparisons across neighborhoods.
+## Data Cleaning and Preparation
+The housing data was prepared by filtering out invalid entries and ensuring that all price data was numeric.
+Only neighborhoods with valid, positive home price values were included.
+The crime data was cleaned by removing any missing entries and focusing exclusively on the crime rates.
+Non-relevant columns (i.e., raw crime counts) were removed in favor of normalized rates per 100,000 residents, making comparisons across neighborhoods more meaningful.
 
 **Focus Neighborhoods**
-The neighborhoods with the 10 highest and 10 lowest home prices were selected for focused analysis.
-These neighborhoods represent a broad spectrum of socio-economic conditions and are expected to exhibit varying crime patterns.
+To investigate the potential relationship between socio-economic status and crime, the analysis focuses on the neighborhoods with the 10 highest and the 10 lowest average home prices.
+These neighborhoods represent a wide socio-economic spectrum and are expected to exhibit varying patterns of criminal activity.
+By isolating these groups, we can assess whether wealthier neighborhoods experience lower crime rates and determine which types of crimes are more prevalent in less affluent areas.
 
-Explain if there were similar datasets that could have been used and why they were not.
-If variables were constructed then this should be mentioned, and high-level cleaning aspects of note should be mentioned, but this section should focus on the destination, not the journey.
-It is important to understand what the variables look like by including graphs, and possibly tables, of all observations, along with discussion of those graphs and the other features of these data.
-Summary statistics should also be included, and well as any relationships between the variables.
-If this becomes too detailed, then appendices could be used.
-
-Below is an overview of the housing dataset, presented in Table 1, and visualizations that illustrate key trends in the data.
+## Summary Statistics and Visualisation 
+Below, we provide an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods, followed by visualizations that illustrate trends in crime rates across Toronto from 2014 to 2023.
 
 ```{r}
+#| fig-pos: "H"
+#| message: false
+#| warning: false
+#| error: false
 #| label: tbl-housing-summary
@@
 # Read in the housing data
-housing_cleaned_data <- read_csv("../data/clean_data/housing_cleaned_data.csv")
+housing_cleaned_data <- read_csv(here::here("data/clean_data/housing_cleaned_data.csv"))
 
-# Summary for the top 10 highest and lowest home prices
-housing_summary <- housing_cleaned_data %>%
+# Get the top 10 and bottom 10 housing prices
+top_10 <- housing_cleaned_data %>%
+  arrange(desc(price)) %>%
+  slice(1:10)
+
+bottom_10 <- housing_cleaned_data %>%
+  arrange(price) %>%
+  slice(1:10)
+
+# Summary for the top 10 highest and bottom 10 lowest home prices
+top_10_summary <- top_10 %>%
   summarise(
+    Group = "Top 10",
     Mean_Price = mean(price),
@@
   )
+
+bottom_10_summary <- bottom_10 %>%
+  summarise(
+    Group = "Bottom 10",
+    Mean_Price = mean(price),
+    Median_Price = median(price),
+    Max_Price = max(price),
+    Min_Price = min(price)
+  )
+
+# Combine the summaries for top 10 and bottom 10
+housing_summary <- bind_rows(top_10_summary, bottom_10_summary)
 
@@
   )
+```
+## Crime Rate Changes over Time {#subsection}
+Crime rate trends over the years (2014-2023) were analyzed for each crime type to understand patterns and shifts. This is visualized in the following plots:
 
-housing_summary_gt
-
-```
 ```{r}
-#| label: tbl-crime-summary
-#| tbl-cap: "Table 2: Summary Statistics of Crime Rates (2014-2023)"
+#| fig-pos: "H"
+#| message: false
+#| warning: false
+#| error: false
+#| label: fig-total-crime-trends
+#| fig-cap: "Figure 1: Total Crime Rates from 2014-2023"
 #| echo: false
-
 # Read in the crime data
@@
 
-# Summarise crime data by calculating the mean for each crime rate across all years
-crime_summary <- crime_cleaned_data %>%
-  summarise(
-    Mean_Assault_Rate = rowMeans(select(., starts_with("assault_rate")), na.rm = TRUE),
-    Mean_Autotheft_Rate = rowMeans(select(., starts_with("autotheft_rate")), na.rm = TRUE),
-    Mean_Biketheft_Rate = rowMeans(select(., starts_with("biketheft_rate")), na.rm = TRUE),
-    Mean_Breakenter_Rate = rowMeans(select(., starts_with("breakenter_rate")), na.rm = TRUE),
-    Mean_Homicide_Rate = rowMeans(select(., starts_with("homicide_rate")), na.rm = TRUE),
-    Mean_Robbery_Rate = rowMeans(select(., starts_with("robbery_rate")), na.rm = TRUE),
-    Mean_Shooting_Rate = rowMeans(select(., starts_with("shooting_rate")), na.rm = TRUE),
-    Mean_Theftfrommv_Rate = rowMeans(select(., starts_with("theftfrommv_rate")), na.rm = TRUE),
-    Mean_Theftover_Rate = rowMeans(select(., starts_with("theftover_rate")), na.rm = TRUE)
-  )
+# Summarize the mean crime rate for each year by averaging the rates across all crime types
+mean_crime_rates <- crime_cleaned_data %>%
+  group_by(year) %>%
+  summarise(mean_crime_rate = mean(rate, na.rm = TRUE))
 
-# Display the summary table using gt
-crime_summary_gt <- crime_summary %>%
-  gt() %>%
-  tab_header(
-    title = md("**Summary Statistics of Crime Rates (2014-2023)**")
-  )
-
-crime_summary_gt
-
-```
-
-**Discussion of Data Selection**
-I chose this dataset over the census data because...
-# Results {#sec-results}
-# Crime Rate Changes over Time {#subsection}
-Crime rate trends over the years (2014-2023) were analyzed for each crime type to understand patterns and shifts. This is visualized in the following plots:
-```{r}
-#| label: fig-transport-vs-stationary
-#| fig-cap: "Figure 1: Time-series visualization of Crime Rates from 2014-2023"
-#| echo: false
-
-crime_cleaned_data_long <- crime_cleaned_data %>%
-  pivot_longer(
-    cols = contains("rate"),  
-    names_to = "Crime_Type",   
-    values_to = "Rate"         
-  ) %>%
-  separate(Crime_Type, into = c("Crime", "Year"), sep = "_rate_")
-
-# Plot the time-series visualization of crime rates from 2014-2023
-crime_cleaned_data_long %>%
-  ggplot(aes(x = Year, y = Rate, color = Crime)) +
-  geom_line() +
-  facet_wrap(~ Crime, scales = "free_y") +
-  labs(title = "Crime Rate Trends in Toronto (2014-2023)",
+# Create a bar plot of mean crime rates from 2014-2023
+ggplot(mean_crime_rates, aes(x = factor(year), y = mean_crime_rate)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Mean Crime Rates in Toronto (2014-2023)",
        x = "Year",
-       y = "Crime Rate per 100,000 Residents") +
-  theme_minimal()
+       y = "Mean Crime Rate per 100,000 Residents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
+  scale_y_continuous(labels = comma)  # Format y-axis with whole numbers and commas
 
@@
 ```{r}
-#| label: fig-percentile-crime-rates
-#| fig-cap: "Figure 2: Box Plots comparing Crime Rates in Top 10 vs Bottom 10 House Price Neighbourhoods"
+#| fig-pos: "H"
+#| message: false
+#| warning: false
+#| error: false
+#| label: tbl-crime-summary
+#| tbl-cap: "Figure 2: Summary Statistics of Crime Rates (2014-2023)"
 #| echo: false
 
-# Check column names of housing_cleaned_data
-names(housing_cleaned_data)
+# Summarize crime data by calculating the mean rate for each crime type and year
+crime_summary <- crime_cleaned_data %>%
+  group_by(crime, year) %>%
+  summarise(Mean_Rate = mean(rate, na.rm = TRUE)) %>%
+  ungroup()
 
-# Check column names of crime_cleaned_data
-names(crime_cleaned_data)
+# Create a bar plot for each crime type by year
+ggplot(crime_summary, aes(x = factor(year), y = Mean_Rate, fill = crime)) +
+  geom_bar(stat = "identity", position = "dodge") +
+  facet_wrap(~ crime, scales = "free_y") +  # Facet by crime type, allow different y scales
+  labs(title = "Mean Crime Rates by Year (2014-2023) for Each Crime Type",
+       x = "Year",
+       y = "Mean Crime Rate per 100,000 Residents",
+       fill = "Crime Type") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
+```
+# Results {#sec-results}
+
+## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods over 2014 to 2023 for Individual Crimes {#subsection}
+```{r}
+#| fig-pos: "H"
+#| message: false
+#| warning: false
+#| error: false
+#| label: fig-crime-comparison
+#| fig-cap: "Figure 3: Box Plots comparing Crime Rates in Top 10 vs Bottom 10 Home Price Neighborhoods"
+#| echo: false
 
@@
 
-# Add 'neighbourhood_type' to label the neighbourhoods
+# Add 'neighbourhood_type' to label the neighborhoods
 top_10_highest <- top_10_highest %>%
@@
 
-# Filter the crime data for those top 10 highest and lowest priced neighborhoods
+# Merge the crime data with the top-bottom neighborhood data
 crime_high_low <- crime_cleaned_data %>%
-  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code)  # Make sure 'neighbourhood_code' is used here
+  inner_join(top_bottom_neighborhoods, by = c("hood_id" = "neighbourhood_code"))
 
-# Join the filtered crime data with the top/bottom housing data
-# Create an overall crime rate by summing all crime rate columns
-crime_high_low <- crime_cleaned_data %>%
-  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code) %>%
-  mutate(overall_crime_rate = rowSums(select(., contains("rate")), na.rm = TRUE))  # Sum of all crime rates
+# Filter out rows with non-finite or missing crime rates
+crime_high_low <- crime_high_low %>%
+  filter(is.finite(rate) & !is.na(rate))  # Remove non-finite and missing values
 
-# Join with housing data to get neighborhood type labels
-crime_high_low <- crime_high_low %>%
-  left_join(top_bottom_neighborhoods, by = c("hood_id" = "neighbourhood_code"))
-
-# Box plot comparing overall crime rates between high and low-priced neighborhoods
-ggplot(crime_high_low, aes(x = neighbourhood_type, y = overall_crime_rate, fill = neighbourhood_type)) +
+# Create the boxplot
+ggplot(crime_high_low, aes(x = neighbourhood_type, y = rate, fill = neighbourhood_type)) +
   geom_boxplot() +
-  labs(title = "Overall Crime Rate Comparison Between High and Low-Priced Neighborhoods",
+  facet_wrap(~ crime, scales = "free_y") +  # Separate plot for each crime type
+  labs(title = "Average Crime Rate Comparison Between High and Low-Priced Neighborhoods from 2014 to 2023",
        x = "Neighborhood Type",
-       y = "Overall Crime Rate per 100,000 Residents") +
+       y = "Crime Rate per 100,000 Residents") +
   theme_minimal()
@@
 Box Plots Comparing Crime Rates for High vs. Low Price Neighborhoods:
+This plot shows a comparative analysis of crime rates between the Top 10 Highest and Bottom 10 Lowest home price neighborhoods in Toronto from 2014 to 2023.
+Each box plot represents the distribution of crime rates for different types of crime (e.g., assault, autotheft, biketheft) in these neighborhoods, with the y-axis representing the crime rate per 100,000 residents.
+
+Key Observations:
+Assault Rates:
+Assault rates are notably higher in low-priced neighborhoods than in high-priced ones.
+The median assault rate in the bottom 10 neighborhoods is around 500, while in the top 10 neighborhoods, it is closer to 250.
+The top neighborhoods also show less variability in assault rates.
+Auto Theft:
+Auto theft rates are slightly higher in bottom-priced neighborhoods, but the difference between the two groups is not as pronounced as for assault.
+Both groups exhibit a broad distribution, with several outliers in low-priced neighborhoods indicating significantly higher rates in certain years.
+Bike Theft:
+High-priced neighborhoods have a notably higher rate of bike theft compared to low-priced ones.
+The median bike theft rate in top neighborhoods is significantly larger, suggesting that wealthier neighborhoods may experience more of these types of non-violent crimes, possibly due to the greater number of bikes or higher-value targets.
+Break and Enter:
+The break and enter crime rate appears to be higher in low-priced neighborhoods, with a median crime rate around 400 incidents per 100,000 residents, compared to around 250 in high-priced neighborhoods.
+The spread of crime rates is more pronounced in low-priced areas.
+Homicide:
+Homicide rates in low-priced neighborhoods are significantly higher than in high-priced ones.
+While most of the values cluster around low rates (under 10 per 100,000 residents), there are some extreme outliers in low-priced neighborhoods.
+Robbery:
+Robbery rates are consistently higher in low-priced neighborhoods compared to high-priced ones.
+The difference in medians is noticeable, with the bottom 10 neighborhoods showing a broader range of robbery rates.
+Shooting:
+Shootings are more common in low-priced neighborhoods, with significantly higher variability and outliers.
+In contrast, high-priced neighborhoods show a relatively small number of shootings, with rates clustering below 25 incidents per 100,000 residents.
+Theft from Motor Vehicles:
+Theft from motor vehicles (theftfrommv) shows similar crime rates between both neighborhood groups. However, there is a slightly higher median and more variability in bottom-priced neighborhoods.
+Theft Over $5,000:
+Both groups exhibit similar trends in "theft over $5,000" rates. There is not a significant difference between the two groups, and both have fairly consistent rates over the years.
+Insights:
+Higher Crime in Low-Priced Neighborhoods: Violent crimes like assault, break and enter, homicide, robbery, and shooting tend to be more prevalent in neighborhoods with lower home prices.
+These areas also exhibit greater variability in crime rates, possibly reflecting more social instability or economic distress.
+Higher Non-Violent Crimes in High-Priced Neighborhoods: Interestingly, non-violent crimes like bike theft are more common in higher-priced neighborhoods, possibly reflecting the availability of more valuable goods in these areas.
+Socioeconomic Disparities in Crime: This analysis confirms the hypothesis that wealthier neighborhoods experience lower rates of violent crimes, while economically disadvantaged areas face greater exposure to violent crime.
+However, non-violent crimes (e.g., thefts) seem more prevalent in wealthier neighborhoods, suggesting different socio-economic dynamics at play.
 
@@
 
-```{r}
-
-```
-
-Bar Plots Comparing Crime Rates Between High and Low Price Neighborhoods:
-This will provide a clear comparison between the two groups for each type of crime.
+## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods from 2014 to 2023 for Overall Crime {#subsection}
 
 ```{r}
-#| label: fig-transport-vs-stationary
-#| fig-cap: "Figure 2: Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
+#| fig-pos: "H"
+#| message: false
+#| warning: false
+#| error: false
+#| label: fig-crime-time-series
+#| fig-cap: "Figure 4: Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
 #| echo: false
 
-# Bar plot comparing crime rates for top 10 highest vs. lowest price neighborhoods
-# First, pivot the crime data longer to create a year column
-crime_high_low_long <- crime_cleaned_data %>%
-  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code) %>%
-  pivot_longer(cols = contains("rate"), 
-               names_to = "Crime_Type_Year", 
-               values_to = "Rate")
+# Filter the crime data for top 10 highest and lowest priced neighborhoods
+crime_high_low <- crime_cleaned_data %>%
+  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code)
 
-# Separate the Crime_Type_Year column into 'Crime_Type' and 'Year'
-crime_high_low_long <- crime_high_low_long %>%
-  separate(Crime_Type_Year, into = c("Crime_Type", "Year"), sep = "_rate_")
-
-# Convert the 'Year' column to numeric
-crime_high_low_long <- crime_high_low_long %>%
-  mutate(Year = as.numeric(Year))
-
-# Calculate the overall crime rate for each neighborhood and year
-# Using either rowSums (for summing) or rowMeans (for averaging)
-crime_high_low_yearly <- crime_high_low_long %>%
-  group_by(hood_id, Year) %>%
-  summarise(overall_crime_rate = sum(Rate, na.rm = TRUE)) %>%
+# Calculate the overall crime rate by summing the rates of all crime types per neighborhood and year
+crime_high_low_yearly <- crime_high_low %>%
+  group_by(hood_id, year) %>%
+  summarise(overall_crime_rate = sum(rate, na.rm = TRUE)) %>%
   ungroup()
 
-# Join with housing data to get neighborhood type labels
+# Join the crime data with the top-bottom neighborhood data to get neighborhood type labels
 crime_high_low_yearly <- crime_high_low_yearly %>%
@@
 crime_yearly_summary <- crime_high_low_yearly %>%
-  group_by(neighbourhood_type, Year) %>%
+  group_by(neighbourhood_type, year) %>%
   summarise(avg_overall_crime_rate = mean(overall_crime_rate, na.rm = TRUE)) %>%
@@
 
-# Bar plot comparing overall crime rates between top and bottom-priced neighborhoods for each year
-ggplot(crime_yearly_summary, aes(x = Year, y = avg_overall_crime_rate, fill = neighbourhood_type)) +
+# Create the bar plot comparing overall crime rates between top and bottom-priced neighborhoods for each year
+ggplot(crime_yearly_summary, aes(x = year, y = avg_overall_crime_rate, fill = neighbourhood_type)) +
   geom_bar(stat = "identity", position = "dodge") +
@@
 ```
+## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods from 2014 to 2023 for Individual Crimes {#subsection}
 ```{r}
-# First, pivot the crime data longer to create a year column
-crime_high_low_long <- crime_cleaned_data %>%
-  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code) %>%
-  pivot_longer(cols = contains("rate"), 
-               names_to = "Crime_Type_Year", 
-               values_to = "Rate")
+#| message: false
+#| warning: false
+#| error: false
+#| fig-pos: "H"
+#| label: fig-crime-time-series-individuals
+#| fig-cap: "Figure 5: Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
+#| echo: false
 
-# Separate the Crime_Type_Year column into 'Crime_Type' and 'Year'
-crime_high_low_long <- crime_high_low_long %>%
-  separate(Crime_Type_Year, into = c("Crime_Type", "Year"), sep = "_rate_")
+# List of crimes to investigate
+crimes_to_investigate <- c("assault", "autotheft", "biketheft", "breakenter", "homicide", "robbery", "shooting", "theftfrommv", "theftover")
 
-# Convert the 'Year' column to numeric
-crime_high_low_long <- crime_high_low_long %>%
-  mutate(Year = as.numeric(Year))
+# Filter for top/bottom 10 neighborhoods
+crime_high_low <- crime_cleaned_data %>%
+  filter(hood_id %in% top_bottom_neighborhoods$neighbourhood_code)
 
-# Join with housing data to get neighborhood type labels
-crime_high_low_long <- crime_high_low_long %>%
+# Join the crime data with the top-bottom neighborhood data to get neighborhood type labels
+crime_high_low <- crime_high_low %>%
   left_join(top_bottom_neighborhoods, by = c("hood_id" = "neighbourhood_code"))
-### Bar Plot for Assault Rates ###
 
-# Filter for assault rates
-assault_rates <- crime_high_low_long %>%
-  filter(Crime_Type == "assault") %>%
-  group_by(neighbourhood_type, Year) %>%
-  summarise(avg_assault_rate = mean(Rate, na.rm = TRUE)) %>%
-  ungroup()
+# Group by crime type, neighborhood type, and year to calculate the average crime rate
+crime_high_low_summary <- crime_high_low %>%
+  group_by(crime, neighbourhood_type, year) %>%
+  summarise(avg_crime_rate = mean(rate, na.rm = TRUE), .groups = 'drop')
 
-# Create bar plot for assault rates
-ggplot(assault_rates, aes(x = Year, y = avg_assault_rate, fill = neighbourhood_type)) +
+# Create a single bar plot with facets for each crime type
+ggplot(crime_high_low_summary, aes(x = year, y = avg_crime_rate, fill = neighbourhood_type)) +
   geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Assault Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
+  facet_wrap(~ crime, scales = "free_y") +  # Facet for each crime type
+  labs(title = "Crime Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
        x = "Year",
-       y = "Average Assault Rate per 100,000 Residents") +
+       y = "Average Crime Rate per 100,000 Residents",
+       fill = "Neighborhood Type") +
   theme_minimal() +
-  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))
-```
-
-```{r}
-
-### Bar Plot for Motor Vehicle Theft Rates (Autotheft) ###
-
-# Filter for motor vehicle theft (autotheft) rates
-autotheft_rates <- crime_high_low_long %>%
-  filter(Crime_Type == "autotheft") %>%
-  group_by(neighbourhood_type, Year) %>%
-  summarise(avg_autotheft_rate = mean(Rate, na.rm = TRUE)) %>%
-  ungroup()
-
-# Create bar plot for motor vehicle theft rates
-ggplot(autotheft_rates, aes(x = Year, y = avg_autotheft_rate, fill = neighbourhood_type)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Motor Vehicle Theft Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
-       x = "Year",
-       y = "Average Motor Vehicle Theft Rate per 100,000 Residents") +
-  theme_minimal() +
-  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))
-```
-```{r}
-### Bar Plot for Shooting Rates ###
-
-# Filter for shooting rates
-assault_rates <- crime_high_low_long %>%
-  filter(Crime_Type == "shooting") %>%
-  group_by(neighbourhood_type, Year) %>%
-  summarise(avg_assault_rate = mean(Rate, na.rm = TRUE)) %>%
-  ungroup()
-
-# Create bar plot for assault rates
-ggplot(assault_rates, aes(x = Year, y = avg_assault_rate, fill = neighbourhood_type)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Shooting Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
-       x = "Year",
-       y = "Average Shooting Rate per 100,000 Residents") +
-  theme_minimal() +
-  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))
+  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red")) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
 
 ```
-```{r}
-### Bar Plot for Break and Enter Rates ###
-
-# Filter for shooting rates
-assault_rates <- crime_high_low_long %>%
-  filter(Crime_Type == "breakenter") %>%
-  group_by(neighbourhood_type, Year) %>%
-  summarise(avg_assault_rate = mean(Rate, na.rm = TRUE)) %>%
-  ungroup()
-
-# Create bar plot for assault rates
-ggplot(assault_rates, aes(x = Year, y = avg_assault_rate, fill = neighbourhood_type)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Break and Enter Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
-       x = "Year",
-       y = "Average Break and Enter Rate per 100,000 Residents") +
-  theme_minimal() +
-  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))
-
-```
-```{r}
-### Bar Plot for Robbery Rates ###
-
-# Filter for shooting rates
-assault_rates <- crime_high_low_long %>%
-  filter(Crime_Type == "robbery") %>%
-  group_by(neighbourhood_type, Year) %>%
-  summarise(avg_assault_rate = mean(Rate, na.rm = TRUE)) %>%
-  ungroup()
-
-# Create bar plot for assault rates
-ggplot(assault_rates, aes(x = Year, y = avg_assault_rate, fill = neighbourhood_type)) +
-  geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Robbery Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
-       x = "Year",
-       y = "Average Robbery Rate per 100,000 Residents") +
-  theme_minimal() +
-  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))
-
-```
-Figure 2 breaks down the number of clients transported to shelters versus those who remained stationary on buses at Spadina Station, providing a deeper understanding of how the initiative's resources were utilized.
-
-The number of
-
-# Model
-
-The goal of our modeling strategy is twofold.
-Firstly, we aim to understand the factors driving the demand for transportation to warming centers versus the decision to remain stationary.
-Here, we briefly describe the Bayesian analysis model used to investigate these patterns.
-Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of clients using the service on day $i$. We model the relationship between $y_i$ and weather conditions using a Bayesian regression model:
-
-\begin{align} y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \ \mu_i &= \alpha + \beta_1 \times \text{temperature}_i + \beta_2 \times \text{precipitation}_i \ \alpha &\sim \mbox{Normal}(0, 2.5) \ \beta_1, \beta_2 &\sim \mbox{Normal}(0, 2.5) \ \sigma &\sim \mbox{Exponential}(1) \end{align}
-
-We run the model in R [@citeR] using the rstanarm package from @rstanarm. We use the default priors from rstanarm.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-### Model justification
-
-We expect a positive relationship between adverse weather conditions (e.g., lower temperatures and higher precipitation) and the number of clients using the service.
-This is based on the assumption that more individuals will seek shelter during harsher weather conditions.
+Crime-Specific Analysis:
+Assault:
+The assault rates remain consistently higher in Bottom 10 Lowest neighborhoods than in Top 10 Highest neighborhoods across all years.
+There seems to be a slight downward trend in the top neighborhoods around 2019–2021, while rates in lower neighborhoods remain relatively stable.
+Autotheft:
+Bottom 10 Lowest neighborhoods show a steady increase in autotheft rates from 2015 onward, peaking around 2022.
+Top 10 Highest neighborhoods have significantly lower autotheft rates throughout the period, but they also show a slight upward trend, though the increase is far more gradual.
+Biketheft:
+In the earlier years (2015–2017), Bottom 10 Lowest neighborhoods have higher bike theft rates, but Top 10 Highest neighborhoods catch up around 2018–2020.
+From 2020 onwards, the bike theft rates in Top 10 Highest neighborhoods are actually higher than those in the Bottom 10 Lowest neighborhoods.
+Breakenter (Breaking and Entering):
+Rates for break and enter crimes fluctuate more in Bottom 10 Lowest neighborhoods, peaking around 2016–2017 and then decreasing slightly.
+Top 10 Highest neighborhoods show more stable but much lower break and enter rates overall.
+Homicide:
+Both neighborhoods show very low homicide rates (as expected), but Bottom 10 Lowest neighborhoods exhibit slightly higher rates.
+There is a gradual upward trend in homicides in Bottom 10 Lowest neighborhoods from 2016 to 2023, although the scale is relatively small (0 to 7.5 per 100,000 residents).
+Robbery:
+Robbery rates in Bottom 10 Lowest neighborhoods are consistently higher than in Top 10 Highest neighborhoods.
+The gap between the two neighborhood types widens in 2020 and onwards, with a notable increase in robbery rates in Bottom 10 Lowest neighborhoods.
+Shooting:
+Bottom 10 Lowest neighborhoods experience a significant peak in shooting rates around 2018, after which they decline.
+Top 10 Highest neighborhoods have almost negligible shooting rates in comparison.
+Theftfrommv (Theft from Motor Vehicle):
+The theft from motor vehicle rates are much higher in Bottom 10 Lowest neighborhoods, and they fluctuate over the years, with no consistent trend.
+The rates for Top 10 Highest neighborhoods remain fairly stable and low across the entire time period.
+Theftover (Theft over $5,000):
+Theft over $5,000 shows stable but relatively low rates for both neighborhood types, though Bottom 10 Lowest neighborhoods have a slight edge.
+There's an increasing trend in Bottom 10 Lowest neighborhoods starting in 2020.
 
@@
 Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory Models of Client Usage Based on Weather Conditions"
-#| warning: false
-
-library(rstanarm)
-
-first_model <- readRDS(file = here::here("models/first_model.rds"))
-
-modelsummary::modelsummary(
-  list(
-    "First Model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
 
@@
 
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
 ```
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
 \newpage

### 2024-09-20 scripts/02-data_cleaning.R

@@ -47,12 +47,22 @@ raw_data_crime <- read_csv("./data/raw_data/crime_raw_data.csv")
 # Cleaning steps for crime data:
 crime_cleaned_data <- raw_data_crime %>%
   janitor::clean_names() %>%
-  # Select columns of neighborhood id and columns relevant to different crime rates
+  # Select neighborhood ID and crime rate columns
   select("hood_id", starts_with("assault_rate"), starts_with("autotheft_rate"),
          starts_with("biketheft_rate"), starts_with("breakenter_rate"),
          starts_with("homicide_rate"), starts_with("robbery_rate"),
          starts_with("shooting_rate"), starts_with("theftfrommv_rate"),
-         starts_with("theftover_rate")) 
+         starts_with("theftover_rate")) %>%
+  # Pivot to long format to extract year and crime type
+  pivot_longer(
+    cols = -hood_id,  # Keep neighborhood ID intact
+    names_to = "crime_year",  # Combine crime type and year
+    values_to = "rate"        # Values are the crime rates
+  ) %>%
+  # Separate the 'crime_year' column into 'crime' and 'year'
+  separate(crime_year, into = c("crime", "year"), sep = "_rate_") %>%
+  # Convert year to numeric 
+  mutate(year = as.numeric(year))
 
 #### Save cleaned crime data ####
 write_csv(crime_cleaned_data, "./data/clean_data/crime_cleaned_data.csv")

### 2024-09-23 scripts/00-simulate_data.R
Cleaned files using stylr and updated simulated data and corresponding tests to better reflect cleaned data.

@@ -1,65 +1,62 @@
 #### Preamble ####
 # Purpose: Simulates crime rate trends using Poisson distribution for the bottom 10 and top 10 housing price neighborhoods over the years 2014 to 2023 and saves the corresponding line plot to analysis_data.
 # Author: Chenika Bukes
-# Date: 20 September 2024
+# Date: 24 September 2024
 # Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
-# - Installed `tidyverse` package
+# Pre-requisites:
+# - Installed `tidyverse` and `purrr` packages
 
 #### Workspace setup ####
 library(tidyverse)
+library(purrr)
 
-#### Simulate data ####
+#### Simulate Data ####
+set.seed(888) 
 
-# Set seed for reproducibility
-set.seed(888)
-
-# Define the years for the simulation period (2014 to 2023)
+# Number of years and crime types
 years <- 2014:2023
 n_years <- length(years)
+crime_types <- c("assault", "autotheft", "biketheft", "breakenter", "homicide", "robbery", "shooting", "theftfrommv", "theftover")
+n_crime_types <- length(crime_types)
 
-# Create the grid of all combinations of neighborhoods and years
-data_grid <- expand.grid(
-  Neighborhood = c(rep("Bottom 10", 10), rep("Top 10", 10)),
+# Simulate house prices for top 10 and bottom 10 neighborhoods
+n_neighborhoods <- 10
+
+# Average house prices for top 10 and bottom 10
+top_10_avg_price <- 1500000
+bottom_10_avg_price <- 400000
+
+# Simulate house prices with some noise
+top_10_prices <- rnorm(n_neighborhoods, mean = top_10_avg_price, sd = 100000)
+bottom_10_prices <- rnorm(n_neighborhoods, mean = bottom_10_avg_price, sd = 50000)
+
+# Create data frame for house prices
+house_prices <- data.frame(
+  Neighborhood = c(paste0("Top_", 1:10), paste0("Bottom_", 1:10)),
+  Price = c(top_10_prices, bottom_10_prices),
+  Group = rep(c("Top 10", "Bottom 10"), each = n_neighborhoods)
+)
+
+# Simulate crime rates for each neighborhood, crime type, and year
+crime_rates <- expand.grid(
+  Neighborhood = c(paste0("Top_", 1:10), paste0("Bottom_", 1:10)),
+  Crime_Type = crime_types,
   Year = years
 )
 
[…]

### 2024-09-23 scripts/01-download_data.R

@@ -4,12 +4,12 @@
 # Date: 20 September 2024
 # Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
+# Pre-requisites:
 #   - Installed the `opendatatoronto` and `tidyverse` packages
 #   - An internet connection with live Open Data Toronto api to download the data
-# Any other information needed? 
+# Any other information needed?
 #   - This script downloads data for the Neighborhood Crime Rates dataset from Open Data Toronto and saves it as a CSV file for further analysis.
-#   - The housing price data was downloaded directly from the site as an excel file. 
+#   - The housing price data was downloaded directly from the site as an excel file.
 
 #### Workspace setup ####
 library(opendatatoronto)
@@ -17,22 +17,18 @@ library(tidyverse)
 
 #### Download Crime Statstic Data for each Neighborhood ####
 # Get the package information for the Neighborhood Crime Rates
-package <- show_package("neighbourhood-crime-rates")  
+package <- show_package("neighbourhood-crime-rates")
 package
 
 # Get all resources for this package
-resources <- list_package_resources("neighbourhood-crime-rates")  
+resources <- list_package_resources("neighbourhood-crime-rates")
 
 # Identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+datastore_resources <- filter(resources, tolower(format) %in% c("csv", "geojson"))
 
 # Load the first datastore resource as a sample
 the_raw_data <- filter(datastore_resources, row_number() == 1) %>% get_resource()
 the_raw_data
 
 #### Save data ####
-write_csv(the_raw_data, "./data/raw_data/crime_raw_data.csv") 
-
-         
-
-
+write_csv(the_raw_data, "./data/raw_data/crime_raw_data.csv")

### 2024-09-23 scripts/02-data_cleaning.R

@@ -1,13 +1,13 @@
 #### Preamble ####
-# Purpose: Cleans the raw housing data 
-# Author: Chenika Bukes 
-# Date: 20 September 2024 
-# Contact: chenika.bukes@mail.utoronto.ca 
+# Purpose: Cleans the raw housing data
+# Author: Chenika Bukes
+# Date: 20 September 2024
+# Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
+# Pre-requisites:
 #   - Installed the `tidyverse` and `janitor` R packages
 #   - Raw housing and crime data files saved as CSV or Excel
-# Any other information needed? 
+# Any other information needed?
 # This script processes and cleans both housing price and crime rate data for further analysis.
 
 #### Workspace setup ####
@@ -20,19 +20,17 @@ library(readxl)
 raw_data_housing <- read_excel("./data/raw_data/wellbeing-toronto-housing.xlsx", sheet = 3)
 
 # Cleaning steps:
-housing_cleaned_data <- 
-  raw_data_housing |> 
-  janitor::clean_names() |> # Clean column names (makes them lower_snake_case)
-  select(neighbourhood, neighbourhood_id, home_prices) |> # Select relevant columns
+housing_cleaned_data <-
+  raw_data_housing |>
+  janitor::clean_names() |>
+  select(neighbourhood, neighbourhood_id, home_prices) |>
   mutate(
-    # Clean and convert "home_prices" to numeric if needed
     home_prices = as.numeric(gsub(",", "", home_prices)) # Remove commas and convert to numeric
-  ) |> 
-  # If any neighborhood has invalid/missing home price, remove the row
-  drop_na(home_prices) |> 
-  filter(home_prices > 0) |>  
+  ) |>
+  drop_na(home_prices) |>
+  filter(home_prices > 0) |>
   rename(
-    neighbourhood_name = neighbourhood, 
+    neighbourhood_name = neighbourhood,
     neighbourhood_code = neighbourhood_id,
     price = home_prices
   )
@@ -48,22 +46,21 @@ raw_data_crime <- read_csv("./data/raw_data/crime_raw_data.csv")
 crime_cleaned_data <- raw_data_crime %>%
   janitor::clean_names() %>%
   # Select neighborhood ID and crime rate columns
-  select("hood_id", starts_with("assault_rate"), starts_with("autotheft_rate"),
-         starts_with("biketheft_rate"), starts_with("breakenter_rate"),
[…]

### 2024-09-23 scripts/03-test_data.R

@@ -1,10 +1,10 @@
 #### Preamble ####
-# Purpose: Test and validate the simulated crime rate data.
+# Purpose: Test and validate the simulated crime rate and home price data.
 # Author: Chenika Bukes
-# Date: 20 September 2024
+# Date: 24 September 2024
 # Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
+# Pre-requisites:
 # - Installed `tidyverse` package
 
 #### Workspace setup ####
@@ -15,16 +15,16 @@ simulated_data <- read_csv("./data/analysis_data/simulated_crime_data.csv")
 
 #### Data validation tests ####
 
-# Test 1: Check that the crime rates are within a reasonable range (0 to 100)
-test_crime_rate_range <- all(simulated_data$Crime_Rate >= 0 & simulated_data$Crime_Rate <= 10000)
+# Test 1: Check that the crime rates are within a reasonable range for Toronto (0 to 1000, as per simulation)
+test_crime_rate_range <- all(simulated_data$Rate >= 0 & simulated_data$Rate <= 1000)
 if (!test_crime_rate_range) {
-  stop("Test 1 failed: Crime rates are outside the expected range (0-100).")
+  stop("Test 1 failed: Crime rates are outside the expected range (0-1000).")
 } else {
   message("Test 1 passed: Crime rates are within the expected range.")
 }
 
 # Test 2: Check that there are no negative crime rates
-test_non_negative_crime <- all(simulated_data$Crime_Rate >= 0)
+test_non_negative_crime <- all(simulated_data$Rate >= 0)
 if (!test_non_negative_crime) {
   stop("Test 2 failed: There are negative crime rates in the data.")
 } else {
@@ -32,16 +32,17 @@ if (!test_non_negative_crime) {
 }
 
 # Test 3: Check that average crime rates are consistently higher for Bottom 10 neighborhoods compared to Top 10
+# as per hypothesis
 average_crime_rates <- simulated_data %>%
-  group_by(Neighborhood) %>%
-  summarise(Average_Crime_Rate = mean(Crime_Rate))
+  group_by(Group) %>%
+  summarise(Average_Crime_Rate = mean(Rate))
 
 bottom_10_avg <- average_crime_rates %>%
-  filter(Neighborhood == "Bottom 10") %>%
+  filter(Group == "Bottom 10") %>%
   summarise(Bottom_10_Avg_Crime = mean(Average_Crime_Rate))
 
 top_10_avg <- average_crime_rates %>%
[…]

### 2024-09-23 paper/paper.qmd
Updated paper file to include full analysis and citations.

@@
     toc-title: "Table of Contents" 
-abstract: "This study analyzes crime rates across Toronto's neighborhoods from 2014 to 2023, focusing on the 10 neighborhoods with the highest home prices and the 10 neighborhoods with the lowest home prices.
-We investigated crime trends across nine categories, including assault, robbery, and autotheft, comparing the average crime rates between high-price and low-price neighborhoods.
-The analysis reveals a significant disparity in crime rates, with lower-priced neighborhoods consistently exhibiting higher crime rates, especially in violent crimes such as shootings and robberies.
-These findings underscore the importance of targeted crime prevention strategies in economically disadvantaged neighborhoods and suggest that housing affordability is closely linked to public safety concerns."
-number-sections: true
+abstract: "This paper analyzes crime rates across Toronto's neighborhoods from 2014 to 2023, focusing on the 10 neighborhoods with the highest home prices and the 10 neighborhoods with the lowest home prices.
+Home price acts as a proxy for socio-economic conditions, with higher home prices suggesting wealthier neighborhoods, and lower prices implying economically disadvantaged areas.
+It investigates crime rate trends across nine categories, comparing the average crime rates between high-price and low-price neighborhoods.
+The analysis reveals a significant disparity in crime rates, with lower-priced neighborhoods consistently exhibiting higher violent crime rates such as shootings, assaults, and murders while high-price neighbourhoods experience higher theft crime rates.
+These findings underscore the importance of targeted crime prevention strategies for varying economic standing neighborhoods and suggests that housing affordability is closely linked to public safety concerns."
 bibliography: references.bib
@@
 library("tidyverse")
-library("gt")
 library("scales")
+library("knitr")
 ```
@@
 
-We run the model in R [@citeR].
-Understanding the relationship between housing prices and crime rates is critical for urban planning and public policy.
-Housing prices are often considered a key indicator of socio-economic status, with higher-priced neighborhoods typically associated with lower crime rates, greater economic stability, and improved social services.
-Conversely, lower-priced neighborhoods are often linked to higher crime rates, economic challenges, and reduced access to resources.
-However, while these patterns may hold in many cases, the relationship between housing prices and crime rates is complex and not universally consistent across all regions.
-This complexity raises the need for more detailed investigation into how these factors interact in specific urban settings, such as Toronto.
+Housing prices are often regarded as a key indicator of socioeconomic status, with higher-priced neighborhoods typically associated with lower crime rates, greater economic stability, and improved access to social services.
+Conversely, lower-priced neighborhoods are frequently linked to higher crime rates, financial instability, and reduced access to critical resources [@Kitchen].
+While these patterns may hold true in many instances, the relationship between housing prices and crime rates is not universally consistent across regions.
+This complexity underscores the need for a more detailed investigation into how these factors interact in specific urban environments, such as Toronto.
 
-Despite numerous studies exploring socio-economic indicators and crime, there remains a gap in understanding how housing prices, particularly in high- and low-income neighborhoods, correlate with different types of crimes over time.
-Previous research often treats these neighborhoods as static entities, overlooking the dynamic nature of crime trends and their fluctuations over time.
-This paper aims to fill this gap by analyzing the crime rates across Toronto’s neighborhoods from 2014 to 2023, specifically focusing on the 10 neighborhoods with the highest and the 10 neighborhoods with the lowest home prices.
+This paper analyzes crime trends and their fluctuations over time across Toronto’s neighborhoods from 2014 to 2023, focusing on the 10 neighborhoods with the highest and lowest average home prices as a proxy for socioeconomic status.
+Using this comparison, I examine the relationship between neighborhood wealth and crime rates to uncover potential socio-economic patterns in crime prevalence.
+The analysis explores nine distinct crime categories and compares trends between high-priced and low-priced neighborhoods.
+The objective is to identify significant crime patterns, assess whether wealthier neighborhoods consistently experience lower crime rates, and investigate which types of crimes are most prevalent in different socio-economic contexts.
 
-In this analysis, we examined nine distinct crime categories—including assault, robbery, and autotheft—and compared the trends in both high- and low-priced neighborhoods.
-The goal was to identify significant patterns in crime rates, understand whether wealthier neighborhoods consistently experience lower crime rates, and investigate which types of crimes are most prevalent in different socio-economic contexts.
-Our findings reveal notable disparities, particularly in violent crimes such as shootings and robberies, which were found to be more prevalent in lower-income neighborhoods.
+Our findings reveal marked disparities, particularly in violent crimes such as shootings and assaults, which are more prevalent in lower home price neighbourhoods.
+As well, non-violent crimes such as theft, had signficantly higher rates in higher home price neighbourhoods.
+Violent crimes (assault, homicide, shootings) and autotheft showed the most significant up trend from 2014 to 2023.
+All crimes stagnated or dropped during 2020-2021 which can be attributed to the Covid-19 pandemic lockdowns.
+These insights are crucial for urban planning and policy development, as they highlight the need for targeted crime prevention strategies.
+By understanding how crime rates vary across neighborhoods with different economic conditions, city planners and policymakers can allocate resources more effectively and design interventions tailored to specific socio-economic groups.
 
-These findings are important because they offer insights that can help city planners and policymakers design targeted crime prevention strategies.
-By understanding how crime rates differ across neighborhoods with varying economic conditions, more effective resource allocation and policy interventions can be implemented.
-The structure of this paper is as follows: first, we review the existing literature on the relationship between housing prices and crime.
-Second, we present our methodology and data sources.
-Third, we provide a detailed analysis of the results, followed by a discussion of the implications for public policy and future research.
+The analysis in this paper was conducted using R [@citeR].
+The paper is structured as follows: first, I review the selected data sources, their relevance to the subject, and how the data was cleaned using the janitor [@janitor], readxl [@readxl], tidyverse [@tidyverse] packages.
+Second, I present key visualizations using the scales [@scales] and knitr [@knitr] packages to explore the relationships between neighborhood home prices, crime rates, and temporal trends from 2014 to 2023.
+Third, I provide an analysis of these visualizations and discuss their potential implications for public policy and future research into neighborhood-specific crime prevention strategies, based on the observed temporal trends in crime rates.
 
@@
 ##  Raw Data {#sub-section}
-The data for this analysis comes from two primary sources obtained from Toronto’s Open Data portal:
+The data for this analysis comes from two primary sources obtained from Toronto’s Open Data portal [@OpenDataPortal]:
 
-**Housing Price Data**: This dataset includes information on average home prices across various neighborhoods in Toronto.
+**Housing Price Data**: This dataset includes information on average home prices across various neighborhoods in Toronto [@WellbeingToronto].
 It serves as an indicator of the socio-economic status of each neighborhood.
 The data is extracted from the "Wellbeing Toronto" dataset, which records various socio-economic variables across the city.
+Wellbeing Toronto is a "website that allows you to learn more about the neighbourhood you work, play or live in" according to the City of Toronto.
+Many City of Toronto divisions, Agencies, Boards and Commissions are involved in gathering this data.
+Wellbeing Toronto only uses municipal information for each of the indicators it publishes.
+They have comprehensive information on how data was gathered for their indicators [@DataGathering] and seem to have taken every precaution for as accurate as possible statistic collection for the neighbourhoods of Toronto.
 For this analysis, the focus is on the average home price for each neighborhood, identified by the neighborhood name and ID.
@@
 **Crime Statistics Data**: This dataset contains detailed annual crime statistics for each neighborhood between 2014 and 2023.
-It includes crime rates per 100,000 residents for a range of criminal activities:
+It was originally published by the Toronto Police Service on the Public Safety Data Portal [@NeighbourhoodCrimeRates].
+The data does not include occurrences of crime that were deemed "unfounded" according to Statistics Canada.
+To determine the rate of a crime, the Toronto Police Service use their police records for crime numbers in each neighbourhood and used data from Environics Analytics for population estimates during the year of crime.
+All of these sources are reputable and it appears thoughtful methods of measurement were applied to most closely capture the population statsitics presented in this dataset.
+It includes crime rates per 100,000 residents for a range of crimes:
+
 - Assault
+
 - Auto theft
+
 - Bike theft
+
 - Break and enter
+
 - Homicide
+
 - Robbery
+
 - Shooting
+
 - Theft from motor vehicle
+
 - Theft over $5,000
+
 The crime data provides a robust view of criminal activity across various categories, offering insight into both violent and non-violent crime trends in different neighborhoods.
 
-**Data Selection Reasoning**
-While Toronto’s census data also contains valuable socio-economic information, the housing price dataset was chosen because it was updated more frequently and provides clearer spatial links between neighborhoods and socio-economic status.
-The census data, recorded in 2015 and 2021, presented two key limitations for this analysis: (1) It was not yet fully updated to 2021 data, and (2) the census dataset organizes data by individual household income rather than neighborhood-level data, making it more challenging to compare neighborhoods directly.
+**Data Selection Reasoning**: While Toronto’s census data also contains valuable socio-economic information, the housing price dataset was chosen because it was updated more frequently and provides clearer spatial links between neighborhoods and socio-economic status.
+The census data, recorded in 2015 and 2021, presented two key limitations for this analysis:
+
+- The features I would have used were not updated to 2021 data, and
+
+- the census dataset organizes data by individual household income rather than neighborhood-level data, making it more challenging to compare neighborhoods directly.
+
 The housing price data, already grouped by neighborhood code, was easier to integrate with the crime statistics for this study's purpose.
 
-**Variables and their Roles**
-The key variables under examination are:
+**Variables and their Roles**: The key variables under examination are:
+
   - Neighborhood Name and ID: Each neighborhood is identified by both a name and a unique ID code.
-  - Home Prices: Average home prices (in CAD) across neighborhoods, reflecting the relative affluence of each area.
-  - Crime Rates: Crime statistics for nine different crime categories (listed above), measured as the number of incidents per 100,000 residents annually.
+
+  - Home Prices: Average home prices across neighborhoods, reflecting the relative affluence of each area.
+
+  - Crime Rates: Crime statistics for nine different crime categories, measured as the number of incidents per 100,000 residents annually.
 These crime rates provide a normalized metric for comparing areas with different population sizes.
@@
 ## Data Cleaning and Preparation
-The housing data was prepared by filtering out invalid entries and ensuring that all price data was numeric.
-Only neighborhoods with valid, positive home price values were included.
-The crime data was cleaned by removing any missing entries and focusing exclusively on the crime rates.
-Non-relevant columns (i.e., raw crime counts) were removed in favor of normalized rates per 100,000 residents, making comparisons across neighborhoods more meaningful.
-
-**Focus Neighborhoods**
+**Housing Price Data**: The housing data was prepared by filtering out invalid entries and ensuring that all price data was numeric.
 To investigate the potential relationship between socio-economic status and crime, the analysis focuses on the neighborhoods with the 10 highest and the 10 lowest average home prices.
+Thus, home price data for all other neighbourhoods were filtred out.
 These neighborhoods represent a wide socio-economic spectrum and are expected to exhibit varying patterns of criminal activity.
@@
 
+**Crime Statistics Data**: The crime data was cleaned by removing any missing entries and focusing exclusively on the crime rates.
+Non-relevant columns (i.e., raw crime counts) were removed in favor of normalized rates per 100,000 residents, making comparisons across neighborhoods more meaningful.
+
 ## Summary Statistics and Visualisation 
-Below, we provide an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods, followed by visualizations that illustrate trends in crime rates across Toronto from 2014 to 2023.
+### Summary Statistics of Housing Price Data 
+Below, we provide an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods.
 
 ```{r}
-#| fig-pos: "H"
+#| tbl-cap: "Summary Statistics of Housing Price Data (Top 10 Highest and Lowest)"
+#| echo: false
 #| message: false
-#| warning: false
-#| error: false
-#| label: tbl-housing-summary
-#| tbl-cap: "Table 1: Summary Statistics of Housing Price Data (Top 10 Highest and Lowest)"
-#| echo: false
+#| show_col_types: FALSE
+#| tbl-pos: "H"
 
@@
     Group = "Bottom 10",
-    Mean_Price = mean(price),
-    Median_Price = median(price),
+    Mean_Price = round(mean(price)),
+    Median_Price = round(median(price)),
     Max_Price = max(price),
@@
 
-# Display the summary table using gt
-housing_summary_gt <- housing_summary %>%
-  gt() %>%
-  tab_header(
-    title = md("**Summary Statistics of Housing Price Data (Top 10 Highest and Lowest)**")
-  )
+# Display the summary table using kable for better PDF rendering
+kable(housing_summary, 
+      caption = "Summary Statistics of Housing Price Data (Top 10 Highest and Lowest)", 
+      align = c('l', 'c', 'c', 'c', 'c'))
+
 ```
-## Crime Rate Changes over Time {#subsection}
+### Mean Crime Rate Changes over Time
 Crime rate trends over the years (2014-2023) were analyzed for each crime type to understand patterns and shifts. This is visualized in the following plots:
@@
 ```{r}
+#| fig-cap: "Total Crime Rates from 2014-2023"
 #| fig-pos: "H"
+#| show_col_types: FALSE
+#| echo: false
 #| message: false
-#| warning: false
-#| error: false
-#| label: fig-total-crime-trends
-#| fig-cap: "Figure 1: Total Crime Rates from 2014-2023"
-#| echo: false
+
 # Read in the crime data
@@
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels for readability
-  scale_y_continuous(labels = comma)  # Format y-axis with whole numbers and commas
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
+  scale_y_continuous(labels = comma)
 
 ```
+Figure 1 showcases a rather steady trend in crime rates with a dip in 2020 and 2021 which could be attributed to Covid-19 lockdowns, followed by a spike in 2023 of approximately 25% increase in mean crime rates across the city of Toronto's neighbourhoods.
 
+### Mean Individual Crime Rate Changes over Time 
 ```{r}
@@
 #| error: false
-#| label: tbl-crime-summary
-#| tbl-cap: "Figure 2: Summary Statistics of Crime Rates (2014-2023)"
+#| fig-cap: "Average Individual Crime Rates over all Neighbourhoods (2014-2023)"
 #| echo: false
+#| fig-width: 7
+#| fig-height: 5
 
@@
   geom_bar(stat = "identity", position = "dodge") +
-  facet_wrap(~ crime, scales = "free_y") +  # Facet by crime type, allow different y scales
+  facet_wrap(~ crime, scales = "free_y") +  
   labs(title = "Mean Crime Rates by Year (2014-2023) for Each Crime Type",
@@
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
+  theme(axis.text.x = element_text(angle = 60, hjust = 1, vjust = 1, size = 8)) +  
+  theme(panel.spacing = unit(2, "lines"))  
 ```
+Figure 2 displays the mean crime rate for each crime type individually.
+There is a clear pattern of crime rates being at their lowest levels during 2020-2021 for all crime types, which can be argued to be due to Covid-19 lockdowns.
+
+**Up-trending Individual Crime Rates**
+
+- Autotheft displays an exponential trend, increasing drastically from 2015 levels.
+
+- Homicide was stagnant until the sudden increase 300% from 2 to 6 in 2023.
+
+- Assaults have increased by 34% since 2020 to 2023.
+
+- Theftover $5000 has also increased by 22% from 2019 levels to 2023.
+
+- Shooting has increased by over 300% since 2015 levels
+
+**Down-trending or Stagnant Individual Crime Rates**
+
+- Theft from motor vehicles rates has remained around 300 per 100,000 from 2015 to 2023.
+
+- Robbery rates have decreased significantly from 2015 to 2023 by about 34%.
+
+- Bike theft rates has decreased from highs in 2017-2019 of 150 to 2023 lows of 100.
+
+- Break and enter rates have remained relatively stagnant from 2015, with dips corresponding to Covid-19 lockdowns.
+
+Overall, there is a clear up-trend in violent crimes and down-trend in "pettier", less-violent crimes.
+
 # Results {#sec-results}
 
-## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods over 2014 to 2023 for Individual Crimes {#subsection}
+## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods for Individual Crimes {#results-1}
 ```{r}
@@
 #| label: fig-crime-comparison
-#| fig-cap: "Figure 3: Box Plots comparing Crime Rates in Top 10 vs Bottom 10 Home Price Neighborhoods"
+#| fig-cap: "Box Plots comparing Crime Rates in Top 10 vs Bottom 10 Home Price Neighborhoods"
 #| echo: false
+#| fig-width: 7
+#| fig-height: 5
 
@@
 top_10_highest <- top_10_highest %>%
-  mutate(neighbourhood_type = "Top 10 Highest")
+  mutate(neighbourhood_type = "Top 10")
 
 top_10_lowest <- top_10_lowest %>%
-  mutate(neighbourhood_type = "Bottom 10 Lowest")
+  mutate(neighbourhood_type = "Bottom 10")
 
@@
        y = "Crime Rate per 100,000 Residents") +
-  theme_minimal()
+  theme_minimal() +
+  theme(
+    plot.title = element_text(size = 10)  
+  )
 
@@
 
-Box Plots Comparing Crime Rates for High vs. Low Price Neighborhoods:
-This plot shows a comparative analysis of crime rates between the Top 10 Highest and Bottom 10 Lowest home price neighborhoods in Toronto from 2014 to 2023.
-Each box plot represents the distribution of crime rates for different types of crime (e.g., assault, autotheft, biketheft) in these neighborhoods, with the y-axis representing the crime rate per 100,000 residents.
+Figure 3 shows a comparative analysis of crime rates between the top 10 highest and bottom 10 lowest home price neighborhoods in Toronto from 2014 to 2023.
 
-Key Observations:
-Assault Rates:
-Assault rates are notably higher in low-priced neighborhoods than in high-priced ones.
+**Key Observations**:
+
+- **Assault**: Assault rates are notably higher in low-priced neighborhoods than in high-priced ones.
 The median assault rate in the bottom 10 neighborhoods is around 500, while in the top 10 neighborhoods, it is closer to 250.
 The top neighborhoods also show less variability in assault rates.
-Auto Theft:
-Auto theft rates are slightly higher in bottom-priced neighborhoods, but the difference between the two groups is not as pronounced as for assault.
+
+- **Autotheft**: Auto theft rates are slightly higher in bottom-priced neighborhoods, but the difference between the two groups is not as pronounced as for assault.
 Both groups exhibit a broad distribution, with several outliers in low-priced neighborhoods indicating significantly higher rates in certain years.
-Bike Theft:
-High-priced neighborhoods have a notably higher rate of bike theft compared to low-priced ones.
+
+- **Biketheft**: High-priced neighborhoods have a notably higher rate of bike theft compared to low-priced ones.
 The median bike theft rate in top neighborhoods is significantly larger, suggesting that wealthier neighborhoods may experience more of these types of non-violent crimes, possibly due to the greater number of bikes or higher-value targets.
-Break and Enter:
-The break and enter crime rate appears to be higher in low-priced neighborhoods, with a median crime rate around 400 incidents per 100,000 residents, compared to around 250 in high-priced neighborhoods.
+
+- **Break and Enter**: The break and enter crime rate appears to be higher in low-priced neighborhoods, with a median crime rate around 400 incidents per 100,000 residents, compared to around 250 in high-priced neighborhoods.
 The spread of crime rates is more pronounced in low-priced areas.
-Homicide:
-Homicide rates in low-priced neighborhoods are significantly higher than in high-priced ones.
+
+- **Homicide**: Homicide rates in low-priced neighborhoods are significantly higher than in high-priced ones.
 While most of the values cluster around low rates (under 10 per 100,000 residents), there are some extreme outliers in low-priced neighborhoods.
-Robbery:
-Robbery rates are consistently higher in low-priced neighborhoods compared to high-priced ones.
+
+- **Robbery**: Robbery rates are consistently higher in low-priced neighborhoods compared to high-priced ones.
 The difference in medians is noticeable, with the bottom 10 neighborhoods showing a broader range of robbery rates.
-Shooting:
-Shootings are more common in low-priced neighborhoods, with significantly higher variability and outliers.
+
+- **Shooting**: Shootings are more common in low-priced neighborhoods, with significantly higher variability and outliers.
 In contrast, high-priced neighborhoods show a relatively small number of shootings, with rates clustering below 25 incidents per 100,000 residents.
-Theft from Motor Vehicles:
-Theft from motor vehicles (theftfrommv) shows similar crime rates between both neighborhood groups. However, there is a slightly higher median and more variability in bottom-priced neighborhoods.
-Theft Over $5,000:
-Both groups exhibit similar trends in "theft over $5,000" rates. There is not a significant difference between the two groups, and both have fairly consistent rates over the years.
-Insights:
-Higher Crime in Low-Priced Neighborhoods: Violent crimes like assault, break and enter, homicide, robbery, and shooting tend to be more prevalent in neighborhoods with lower home prices.
+
+- **Theft from motor vehicles**: this shows similar crime rates between both neighborhood groups. However, there is a slightly higher median and more variability in bottom-priced neighborhoods.
+
+- **Theft over $5,000**: Both groups exhibit similar trends in "theft over $5,000" rates.
+There is not a significant difference between the two groups, and both have fairly consistent rates over the years.
+
+Altogether, we see that there is higher violent crime in low-priced neighborhoods.
+Violent crimes like assault, break and enter, homicide, robbery, and shooting tend to be more prevalent in neighborhoods with lower home prices.
 These areas also exhibit greater variability in crime rates, possibly reflecting more social instability or economic distress.
-Higher Non-Violent Crimes in High-Priced Neighborhoods: Interestingly, non-violent crimes like bike theft are more common in higher-priced neighborhoods, possibly reflecting the availability of more valuable goods in these areas.
-Socioeconomic Disparities in Crime: This analysis confirms the hypothesis that wealthier neighborhoods experience lower rates of violent crimes, while economically disadvantaged areas face greater exposure to violent crime.
-However, non-violent crimes (e.g., thefts) seem more prevalent in wealthier neighborhoods, suggesting different socio-economic dynamics at play.
+As well, there is higher non-violent Crimes in high-priced neighborhoods.
+Interestingly, non-violent crimes like bike theft are more common in higher-priced neighborhoods, possibly reflecting the availability of more valuable goods in these areas.
 
-This helps visualize how crime rates differ between the top 10 highest and lowest-priced neighborhoods.
-
-## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods from 2014 to 2023 for Overall Crime {#subsection}
+## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods from 2014 to 2023 for Overall Crime {#results-2}
 
@@
 #| label: fig-crime-time-series
-#| fig-cap: "Figure 4: Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
+#| fig-cap: "Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
 #| echo: false
@@
   geom_bar(stat = "identity", position = "dodge") +
-  labs(title = "Overall Crime Rate for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
+  labs(title = "Average Overall Crime Rate for Top 10 vs Bottom 10 Home Price Neighborhoods from 2014-2023",
        x = "Year",
@@
   theme_minimal() +
-  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red"))
+  scale_fill_manual(values = c("Top 10" = "blue", "Bottom 10" = "red")) +
+  theme(
+    plot.title = element_text(size = 10)  
+  )
 
 ```
-## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods from 2014 to 2023 for Individual Crimes {#subsection}
+Figure 4 shows that the overall crime rate has increased at a much higher rate in the top 10 highest price neighbourhoods than in the bottom 10 lowest price neighbourhoods.
+Both the highest and lowest priced neighbourhoods have experienced a steady increase in overall crime rates from 2015 to 2023, with the highest crime rate being that of the top 10 highest price neighbourhoods in 2023.
+
+## Average Crime Rate Comparison Between High and Low-Priced Neighborhoods from 2014 to 2023 for Individual Crimes {#results-3}
 ```{r}
@@
 #| label: fig-crime-time-series-individuals
-#| fig-cap: "Figure 5: Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
+#| fig-cap: "Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
 #| echo: false
+#| fig-width: 7
+#| fig-height: 5
 
@@
   geom_bar(stat = "identity", position = "dodge") +
-  facet_wrap(~ crime, scales = "free_y") +  # Facet for each crime type
+  facet_wrap(~ crime, scales = "free_y") +  
   labs(title = "Crime Rates for Top 10 vs Bottom 10 Home Price Neighborhoods by Year",
@@
   theme_minimal() +
-  scale_fill_manual(values = c("Top 10 Highest" = "blue", "Bottom 10 Lowest" = "red")) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability
+  scale_fill_manual(values = c("Top 10" = "blue", "Bottom 10" = "red")) +  
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  theme(
+    plot.title = element_text(size = 10)  
+  )
 
 ```
-Crime-Specific Analysis:
-Assault:
-The assault rates remain consistently higher in Bottom 10 Lowest neighborhoods than in Top 10 Highest neighborhoods across all years.
+Figure 5 displays the average crime rate amongst the top 10 and bottom 10 priced neighbourhoods for each individual crime over the years 2014-2023.
+
+- **Assault**: The assault rates remain consistently higher in Bottom 10 Lowest neighborhoods than in Top 10 Highest neighborhoods across all years.
 There seems to be a slight downward trend in the top neighborhoods around 2019–2021, while rates in lower neighborhoods remain relatively stable.
-Autotheft:
-Bottom 10 Lowest neighborhoods show a steady increase in autotheft rates from 2015 onward, peaking around 2022.
+
+- **Autotheft**: Bottom 10 Lowest neighborhoods show a steady increase in autotheft rates from 2015 onward, peaking around 2022.
 Top 10 Highest neighborhoods have significantly lower autotheft rates throughout the period, but they also show a slight upward trend, though the increase is far more gradual.
-Biketheft:
-In the earlier years (2015–2017), Bottom 10 Lowest neighborhoods have higher bike theft rates, but Top 10 Highest neighborhoods catch up around 2018–2020.
+
+- **Biketheft**: In the earlier years (2015–2017), Bottom 10 Lowest neighborhoods have higher bike theft rates, but Top 10 Highest neighborhoods catch up around 2018–2020.
 From 2020 onwards, the bike theft rates in Top 10 Highest neighborhoods are actually higher than those in the Bottom 10 Lowest neighborhoods.
-Breakenter (Breaking and Entering):
-Rates for break and enter crimes fluctuate more in Bottom 10 Lowest neighborhoods, peaking around 2016–2017 and then decreasing slightly.
+
+- **Break and Enter**: Rates for break and enter crimes fluctuate more in Bottom 10 Lowest neighborhoods, peaking around 2016–2017 and then decreasing slightly.
 Top 10 Highest neighborhoods show more stable but much lower break and enter rates overall.
-Homicide:
-Both neighborhoods show very low homicide rates (as expected), but Bottom 10 Lowest neighborhoods exhibit slightly higher rates.
+
+- **Homicide**: Both neighborhoods show very low homicide rates (as expected), but Bottom 10 Lowest neighborhoods exhibit slightly higher rates.
 There is a gradual upward trend in homicides in Bottom 10 Lowest neighborhoods from 2016 to 2023, although the scale is relatively small (0 to 7.5 per 100,000 residents).
-Robbery:
-Robbery rates in Bottom 10 Lowest neighborhoods are consistently higher than in Top 10 Highest neighborhoods.
+
+- **Robbery**: Robbery rates in Bottom 10 Lowest neighborhoods are consistently higher than in Top 10 Highest neighborhoods.
 The gap between the two neighborhood types widens in 2020 and onwards, with a notable increase in robbery rates in Bottom 10 Lowest neighborhoods.
-Shooting:
-Bottom 10 Lowest neighborhoods experience a significant peak in shooting rates around 2018, after which they decline.
+
+- **Shooting**: Bottom 10 Lowest neighborhoods experience a significant peak in shooting rates around 2018, after which they decline.
 Top 10 Highest neighborhoods have almost negligible shooting rates in comparison.
-Theftfrommv (Theft from Motor Vehicle):
-The theft from motor vehicle rates are much higher in Bottom 10 Lowest neighborhoods, and they fluctuate over the years, with no consistent trend.
+
+- **Theft from Motor Vehicle**: The theft from motor vehicle rates are much higher in Bottom 10 Lowest neighborhoods, and they fluctuate over the years, with no consistent trend.
 The rates for Top 10 Highest neighborhoods remain fairly stable and low across the entire time period.
-Theftover (Theft over $5,000):
-Theft over $5,000 shows stable but relatively low rates for both neighborhood types, though Bottom 10 Lowest neighborhoods have a slight edge.
+
+- **Theft over $5,000**: This crime shows stable but relatively low rates for both neighborhood types, though Bottom 10 Lowest neighborhoods have a slight edge.
 There's an increasing trend in Bottom 10 Lowest neighborhoods starting in 2020.
-
-# Results
-
-Our results are summarized in @tbl-modelresults.
 
@@
 
-## First discussion point {#sec-first-point}
+This analysis confirms the hypothesis that wealthier neighborhoods experience lower rates of violent crimes, while economically disadvantaged areas face greater exposure to violent crime.
+However, non-violent crimes seem more prevalent in wealthier neighborhoods, suggesting different socio-economic dynamics at play.
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+The analysis of crime rates from 2014 to 2023 across Toronto's neighborhoods, focusing on the top 10 highest and bottom 10 lowest-priced neighborhoods, reveals significant disparities in crime patterns based on socioeconomic status.
+The results highlight a clear trend where violent crimes—such as assault, homicide, and shootings—are more prevalent in lower-priced neighborhoods, suggesting a direct correlation between economic disadvantage and exposure to violence.
+On the other hand, non-violent crimes like thefts (especially bike thefts and theft from motor vehicles) are more common in wealthier areas, reflecting different crime dynamics in these neighborhoods.
 
-## Second discussion point
+Key observations include:
 
-## Third discussion point
+- **Violent Crimes**: Assault, shootings, and homicides show higher rates in low-priced neighborhoods, with noticeable outliers and variability, indicating potentially greater social instability.
 
-## Weaknesses and next steps
+- **Non-Violent Crimes**: Higher-priced neighborhoods experience more theft-related crimes, likely due to the presence of higher-value targets.
 
-Weaknesses and next steps should also be included.
+- **Temporal Trends**: There is a city-wide spike in crime rates in 2023, following a period of lower crime during the 2020–2021 COVID-19 lockdowns, emphasizing the complex interactions between public health crises and crime patterns.
+
+These findings suggest that public policy should be tailored to address the specific needs of different economic groups.
+Low-income neighborhoods require increased focus on violence prevention, while higher-income areas might benefit from enhanced property crime deterrence.
+Moreover, the evidence points to the need for long-term, community-based crime reduction strategies that consider the broader social and economic factors contributing to crime.
+
+Future work should focus on expanding the scope of the analysis to include more socioeconomic indicators, such as employment rates and education levels, to better understand the root causes of crime and inform more effective policy interventions.
 
@@
 
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
 \newpage

### 2024-09-23 paper/paper.qmd
Removed redundant files, updated README, and LLM usage file.

@@
 **Data Selection Reasoning**: While Toronto’s census data also contains valuable socio-economic information, the housing price dataset was chosen because it was updated more frequently and provides clearer spatial links between neighborhoods and socio-economic status.
-The census data, recorded in 2015 and 2021, presented two key limitations for this analysis:
+The census data, recorded in 2014 and 2021, presented two key limitations for this analysis:
 
@@
 
-- Autotheft displays an exponential trend, increasing drastically from 2015 levels.
+- Autotheft displays an exponential trend, increasing drastically from 2014 levels.
 
@@
 
-- Shooting has increased by over 300% since 2015 levels
+- Shooting has increased by over 300% since 2014 levels
 
@@
 
-- Theft from motor vehicles rates has remained around 300 per 100,000 from 2015 to 2023.
+- Theft from motor vehicles rates has remained around 300 per 100,000 from 2014 to 2023.
 
-- Robbery rates have decreased significantly from 2015 to 2023 by about 34%.
+- Robbery rates have decreased significantly from 2014 to 2023 by about 34%.
 
@@
 
-- Break and enter rates have remained relatively stagnant from 2015, with dips corresponding to Covid-19 lockdowns.
+- Break and enter rates have remained relatively stagnant from 2014, with dips corresponding to Covid-19 lockdowns.
 
@@
 Figure 4 shows that the overall crime rate has increased at a much higher rate in the top 10 highest price neighbourhoods than in the bottom 10 lowest price neighbourhoods.
-Both the highest and lowest priced neighbourhoods have experienced a steady increase in overall crime rates from 2015 to 2023, with the highest crime rate being that of the top 10 highest price neighbourhoods in 2023.
+Both the highest and lowest priced neighbourhoods have experienced a steady increase in overall crime rates from 2014 to 2023, with the highest crime rate being that of the top 10 highest price neighbourhoods in 2023.
 
@@
 
-- **Autotheft**: Bottom 10 Lowest neighborhoods show a steady increase in autotheft rates from 2015 onward, peaking around 2022.
+- **Autotheft**: Bottom 10 Lowest neighborhoods show a steady increase in autotheft rates from 2014 onward, peaking around 2022.
 Top 10 Highest neighborhoods have significantly lower autotheft rates throughout the period, but they also show a slight upward trend, though the increase is far more gradual.
 
-- **Biketheft**: In the earlier years (2015–2017), Bottom 10 Lowest neighborhoods have higher bike theft rates, but Top 10 Highest neighborhoods catch up around 2018–2020.
+- **Biketheft**: In the earlier years (2014–2017), Bottom 10 Lowest neighborhoods have higher bike theft rates, but Top 10 Highest neighborhoods catch up around 2018–2020.
 From 2020 onwards, the bike theft rates in Top 10 Highest neighborhoods are actually higher than those in the Bottom 10 Lowest neighborhoods.

### 2024-09-24 paper/paper.qmd
Addressed github comments from issues.

@@
     toc-title: "Table of Contents" 
-abstract: "This paper analyzes crime rates across Toronto's neighborhoods from 2014 to 2023, focusing on the 10 neighborhoods with the highest home prices and the 10 neighborhoods with the lowest home prices.
-Home price acts as a proxy for socio-economic conditions, with higher home prices suggesting wealthier neighborhoods, and lower prices implying economically disadvantaged areas.
-It investigates crime rate trends across nine categories, comparing the average crime rates between high-price and low-price neighborhoods.
+abstract: "This paper analyzes crime rates across Toronto's neighborhoods from 2014 to 2023, focusing on the 10 neighborhoods with the highest home prices and the 10 neighborhoods with the lowest home prices where home price acts as a proxy for socio-economic conditions.
+The paper investigates crime rate trends across nine categories, comparing the average crime rates between high-price and low-price neighborhoods.
 The analysis reveals a significant disparity in crime rates, with lower-priced neighborhoods consistently exhibiting higher violent crime rates such as shootings, assaults, and murders while high-price neighbourhoods experience higher theft crime rates.
-These findings underscore the importance of targeted crime prevention strategies for varying economic standing neighborhoods and suggests that housing affordability is closely linked to public safety concerns."
+These findings underscore the importance of targeted crime prevention strategies for varying economic standing neighborhoods."
 bibliography: references.bib
@@
 
-Our findings reveal marked disparities, particularly in violent crimes such as shootings and assaults, which are more prevalent in lower home price neighbourhoods.
-As well, non-violent crimes such as theft, had signficantly higher rates in higher home price neighbourhoods.
-Violent crimes (assault, homicide, shootings) and autotheft showed the most significant up trend from 2014 to 2023.
-All crimes stagnated or dropped during 2020-2021 which can be attributed to the Covid-19 pandemic lockdowns.
+My findings reveal marked disparities, particularly in violent crimes such as shootings, assaults, and homicides, which are more prevalent in lower-priced neighborhoods.
+In contrast, non-violent crimes, such as theft, show significantly higher rates in higher-priced neighborhoods.
+Crimes like assault, homicide, theft over $5,000, and auto theft showed the most significant upward trends from 2014 to 2023.
+Crime rates either stagnated or dropped during 2020–2021, which can be attributed to the COVID-19 pandemic lockdowns.
 These insights are crucial for urban planning and policy development, as they highlight the need for targeted crime prevention strategies.
-By understanding how crime rates vary across neighborhoods with different economic conditions, city planners and policymakers can allocate resources more effectively and design interventions tailored to specific socio-economic groups.
+By understanding how crime rates vary across neighborhoods with different economic conditions, and identifying which crimes have surged in recent years, city planners and policymakers can allocate resources more effectively and design interventions tailored to specific socioeconomic groups.
 
 The analysis in this paper was conducted using R [@citeR].
-The paper is structured as follows: first, I review the selected data sources, their relevance to the subject, and how the data was cleaned using the janitor [@janitor], readxl [@readxl], tidyverse [@tidyverse] packages.
+The paper is structured as follows: first, I review the selected data sources and their relevance to the subject.
 Second, I present key visualizations using the scales [@scales] and knitr [@knitr] packages to explore the relationships between neighborhood home prices, crime rates, and temporal trends from 2014 to 2023.
@@
 It serves as an indicator of the socio-economic status of each neighborhood.
-The data is extracted from the "Wellbeing Toronto" dataset, which records various socio-economic variables across the city.
+For this analysis, the focus is on the average home price for each neighborhood, identified by the neighborhood name and ID.
+This variable acts as a proxy for socio-economic conditions, with higher home prices suggesting wealthier neighborhoods, and lower prices implying economically disadvantaged areas.
+
+**Housing Price Data Measurement**: The data is extracted from the "Wellbeing Toronto" dataset, which records various socio-economic variables across the city.
 Wellbeing Toronto is a "website that allows you to learn more about the neighbourhood you work, play or live in" according to the City of Toronto.
@@
 Wellbeing Toronto only uses municipal information for each of the indicators it publishes.
-They have comprehensive information on how data was gathered for their indicators [@DataGathering] and seem to have taken every precaution for as accurate as possible statistic collection for the neighbourhoods of Toronto.
-For this analysis, the focus is on the average home price for each neighborhood, identified by the neighborhood name and ID.
-This variable acts as a proxy for socio-economic conditions, with higher home prices suggesting wealthier neighborhoods, and lower prices implying economically disadvantaged areas.
+They have comprehensive information on how data was gathered for each of their indicators and seem to have taken every precaution for as accurate as possible statistic collection for the neighbourhoods of Toronto [@DataGathering].
 
-**Crime Statistics Data**: This dataset contains detailed annual crime statistics for each neighborhood between 2014 and 2023.
-It was originally published by the Toronto Police Service on the Public Safety Data Portal [@NeighbourhoodCrimeRates].
-The data does not include occurrences of crime that were deemed "unfounded" according to Statistics Canada.
-To determine the rate of a crime, the Toronto Police Service use their police records for crime numbers in each neighbourhood and used data from Environics Analytics for population estimates during the year of crime.
-All of these sources are reputable and it appears thoughtful methods of measurement were applied to most closely capture the population statsitics presented in this dataset.
-It includes crime rates per 100,000 residents for a range of crimes:
+**Crime Statistics Data**: This dataset contains detailed annual crime statistics for each neighborhood between 2014 and 2023. It includes crime rates per 100,000 residents for a range of crimes:
 
@@
 
-**Data Selection Reasoning**: While Toronto’s census data also contains valuable socio-economic information, the housing price dataset was chosen because it was updated more frequently and provides clearer spatial links between neighborhoods and socio-economic status.
-The census data, recorded in 2014 and 2021, presented two key limitations for this analysis:
+**Crime Statistics Data Measurement**: It was originally published by the Toronto Police Service on the Public Safety Data Portal [@NeighbourhoodCrimeRates].
+The data does not include occurrences of crime that were deemed "unfounded" according to Statistics Canada.
+To determine the rate of a crime, the Toronto Police Service use their police records for crime numbers in each neighbourhood and used data from Environics Analytics for population estimates during the year of crime.
+All of these sources are reputable and it appears thoughtful methods of measurement were applied to most closely capture the population statsitics presented in this dataset.
 
-- The features I would have used were not updated to 2021 data, and
+## Data Selection Reasoning {#sub-data-section}
 
-- the census dataset organizes data by individual household income rather than neighborhood-level data, making it more challenging to compare neighborhoods directly.
+While Toronto’s census data also contains valuable socio-economic information, the housing price dataset was chosen because it provides clearer spatial links between neighborhoods and socio-economic status.
+The neighbourhood profile census data [@NeighbourhoodProfile], recorded in 2015, presented two key limitations for this analysis:
+
+- Only neighbourhood names, not neighbourhood codes, were provided and the names did not correspond precisely to the names used by the police department  in the crime dataset.
+
+- The census dataset organizes data by individual household income rather than neighborhood-level data, making it more challenging to compare neighborhoods directly.
 
@@
 
-**Variables and their Roles**: The key variables under examination are:
+**Variables and their Roles**
+
+The key variables under examination are:
 
@@
 These crime rates provide a normalized metric for comparing areas with different population sizes.
+
 These variables allow for a detailed analysis of the relationship between socio-economic status (as proxied by home prices) and crime rates, both across crime categories and over time.
-
-## Data Cleaning and Preparation
-**Housing Price Data**: The housing data was prepared by filtering out invalid entries and ensuring that all price data was numeric.
-To investigate the potential relationship between socio-economic status and crime, the analysis focuses on the neighborhoods with the 10 highest and the 10 lowest average home prices.
-Thus, home price data for all other neighbourhoods were filtred out.
-These neighborhoods represent a wide socio-economic spectrum and are expected to exhibit varying patterns of criminal activity.
-By isolating these groups, we can assess whether wealthier neighborhoods experience lower crime rates and determine which types of crimes are more prevalent in less affluent areas.
-
-**Crime Statistics Data**: The crime data was cleaned by removing any missing entries and focusing exclusively on the crime rates.
-Non-relevant columns (i.e., raw crime counts) were removed in favor of normalized rates per 100,000 residents, making comparisons across neighborhoods more meaningful.
 
@@
 ### Summary Statistics of Housing Price Data 
-Below, we provide an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods.
+Table \@ref(tbl-housing-summary) an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods.
 
@@
 #| message: false
+#| table-label: tbl-housing-summary
 #| show_col_types: FALSE
@@
 ### Mean Crime Rate Changes over Time
-Crime rate trends over the years (2014-2023) were analyzed for each crime type to understand patterns and shifts. This is visualized in the following plots:
+Crime rate trends over the years (2014-2023) were analyzed for each crime type to understand patterns and shifts. This is visualized in Figure \@ref(fig-crime-comparison-0):
 
@@
 #| fig-pos: "H"
+#| fig-label: fig-crime-comparison-0
 #| show_col_types: FALSE
@@
 ```
-Figure 1 showcases a rather steady trend in crime rates with a dip in 2020 and 2021 which could be attributed to Covid-19 lockdowns, followed by a spike in 2023 of approximately 25% increase in mean crime rates across the city of Toronto's neighbourhoods.
+Figure \@ref(fig-crime-comparison-0) showcases a rather steady trend in crime rates with a dip in 2020 and 2021 which could be attributed to Covid-19 lockdowns, followed by a spike in 2023 of approximately 25% increase in mean crime rates across the city of Toronto's neighbourhoods.
 
@@
 #| error: false
+#| fig-label: fig-crime-comparison-1
 #| fig-cap: "Average Individual Crime Rates over all Neighbourhoods (2014-2023)"
@@
 ```
-Figure 2 displays the mean crime rate for each crime type individually.
+Figure \@ref(fig-crime-comparison-1) displays the mean crime rate for each crime type individually.
 There is a clear pattern of crime rates being at their lowest levels during 2020-2021 for all crime types, which can be argued to be due to Covid-19 lockdowns.
@@
 #| error: false
-#| label: fig-crime-comparison
+#| fig-label: fig-crime-comparison
 #| fig-cap: "Box Plots comparing Crime Rates in Top 10 vs Bottom 10 Home Price Neighborhoods"
@@
 
-# Filter out rows with non-finite or missing crime rates
-crime_high_low <- crime_high_low %>%
-  filter(is.finite(rate) & !is.na(rate))  # Remove non-finite and missing values
-
 # Create the boxplot
@@
   geom_boxplot() +
-  facet_wrap(~ crime, scales = "free_y") +  # Separate plot for each crime type
+  facet_wrap(~ crime, scales = "free_y") +  
   labs(title = "Average Crime Rate Comparison Between High and Low-Priced Neighborhoods from 2014 to 2023",
@@
        y = "Crime Rate per 100,000 Residents") +
+  scale_fill_manual(values = c("Bottom 10" = "#ff5050", "Top 10" = "steelblue")) + 
   theme_minimal() +
@@
 
-Figure 3 shows a comparative analysis of crime rates between the top 10 highest and bottom 10 lowest home price neighborhoods in Toronto from 2014 to 2023.
+Figure \@ref(fig-crime-comparison) shows a comparative analysis of crime rates between the top 10 highest and bottom 10 lowest home price neighborhoods in Toronto from 2014 to 2023.
 
-**Key Observations**:
+**Crimes more Prevalant in High-Price Neighbourhoods**
 
-- **Assault**: Assault rates are notably higher in low-priced neighborhoods than in high-priced ones.
+- **Biketheft**: The median bike theft rate in top neighborhoods is significantly larger, suggesting that wealthier neighborhoods may experience more of these types of non-violent crimes, possibly due to the greater number of bikes or higher-value targets.
+
+- **Theft over $5,000**: High-priced neighborhoods show almost double the median rate of low-priced neighborhoods for theft over $5,000, with much larger outliers.
+
+- **Autotheft**: Although auto theft rates are slightly higher in low-priced neighborhoods, the difference in the interquartile range is minor.
+High-priced neighborhoods, however, have much larger outliers for auto theft.
+
+- **Break and Enter**: The break and enter crime rate is higher in high-priced neighborhoods, with a median of around 350 incidents per 100,000 residents, compared to approximately 250 in low-priced neighborhoods.
+
+**Crimes more Prevalant in Low-Price Neighbourhoods**
+
+- **Assault**: Assault rates are notably higher in low-priced neighborhoods compared to high-priced ones.
 The median assault rate in the bottom 10 neighborhoods is around 500, while in the top 10 neighborhoods, it is closer to 250.
-The top neighborhoods also show less variability in assault rates.
-
-- **Autotheft**: Auto theft rates are slightly higher in bottom-priced neighborhoods, but the difference between the two groups is not as pronounced as for assault.
-Both groups exhibit a broad distribution, with several outliers in low-priced neighborhoods indicating significantly higher rates in certain years.
-
-- **Biketheft**: High-priced neighborhoods have a notably higher rate of bike theft compared to low-priced ones.
-The median bike theft rate in top neighborhoods is significantly larger, suggesting that wealthier neighborhoods may experience more of these types of non-violent crimes, possibly due to the greater number of bikes or higher-value targets.
-
-- **Break and Enter**: The break and enter crime rate appears to be higher in low-priced neighborhoods, with a median crime rate around 400 incidents per 100,000 residents, compared to around 250 in high-priced neighborhoods.
-The spread of crime rates is more pronounced in low-priced areas.
 
@@
 
-- **Robbery**: Robbery rates are consistently higher in low-priced neighborhoods compared to high-priced ones.
+- **Robbery**: Robbery rates are significantly higher in low-priced neighborhoods compared to high-priced ones.
 The difference in medians is noticeable, with the bottom 10 neighborhoods showing a broader range of robbery rates.
@@
 - **Shooting**: Shootings are more common in low-priced neighborhoods, with significantly higher variability and outliers.
-In contrast, high-priced neighborhoods show a relatively small number of shootings, with rates clustering below 25 incidents per 100,000 residents.
+In contrast, high-priced neighborhoods show a relatively small number of shootings.
 
-- **Theft from motor vehicles**: this shows similar crime rates between both neighborhood groups. However, there is a slightly higher median and more variability in bottom-priced neighborhoods.
+- **Theft from motor vehicles**: Theft from motor vehicles shows similar crime rates across both neighborhood groups.
+However, there is a slightly higher median and more variability in low-priced neighborhoods.
 
-- **Theft over $5,000**: Both groups exhibit similar trends in "theft over $5,000" rates.
-There is not a significant difference between the two groups, and both have fairly consistent rates over the years.
+Overall, violent crimes such as assault, homicide, robbery, and shootings tend to be more prevalent in low-priced neighborhoods, which also exhibit greater variability in crime rates, potentially reflecting more socio-economic instability.
+On the other hand, non-violent crimes like bike theft, auto theft, and break and enter are more common in high-priced neighborhoods, possibly due to the presence of more valuable goods.
 
-Altogether, we see that there is higher violent crime in low-priced neighborhoods.
-Violent crimes like assault, break and enter, homicide, robbery, and shooting tend to be more prevalent in neighborhoods with lower home prices.
-These areas also exhibit greater variability in crime rates, possibly reflecting more social instability or economic distress.
-As well, there is higher non-violent Crimes in high-priced neighborhoods.
-Interestingly, non-violent crimes like bike theft are more common in higher-priced neighborhoods, possibly reflecting the availability of more valuable goods in these areas.
+However, this analysis provides a relatively static understanding of crime rates in Toronto neighborhoods.
+To make effective public safety policy changes, it is crucial to examine the temporal trends of these crime statistics in both high-priced and low-priced neighborhoods.
+Understanding how crime rates evolve over time will inform both immediate and long-term policy decisions.
 
@@
 #| error: false
-#| label: fig-crime-time-series
+#| fig-label: fig-crime-time-series
 #| fig-cap: "Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
@@
   theme_minimal() +
-  scale_fill_manual(values = c("Top 10" = "blue", "Bottom 10" = "red")) +
+  scale_fill_manual(values = c("Top 10" = "steelblue", "Bottom 10" = "#ff5050")) +
   theme(
-    plot.title = element_text(size = 10)  
+    plot.title = element_text(size = 10) ,
+    axis.title.y = element_text(size = 9)
   )
@@
 ```
-Figure 4 shows that the overall crime rate has increased at a much higher rate in the top 10 highest price neighbourhoods than in the bottom 10 lowest price neighbourhoods.
-Both the highest and lowest priced neighbourhoods have experienced a steady increase in overall crime rates from 2014 to 2023, with the highest crime rate being that of the top 10 highest price neighbourhoods in 2023.
+Figure \@ref(fig-crime-time-series) shows that the overall crime rate has increased at a much faster rate in the top 10 highest-priced neighborhoods compared to the bottom 10 lowest-priced neighborhoods.
+The low-priced neighborhoods maintained relatively consistent crime rates around 1,500 per 100,000 residents until 2022, when the rate began to rise, reaching closer to 2,000.
+Both high-priced and low-priced neighborhoods experienced dips in crime rates over the years but show an overall increasing trend from 2014 to 2023.
+The top 10 neighborhoods have seen the most significant surge, with crime rates rising from a low of about 1,100 in 2016 to a high of about 1,950 in 2023, making the top 10 neighborhoods the ones with the highest crime rate in 2023.
 
@@
 #| fig-pos: "H"
-#| label: fig-crime-time-series-individuals
+#| fig-label: fig-crime-time-series-individuals
 #| fig-cap: "Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
@@
   theme_minimal() +
-  scale_fill_manual(values = c("Top 10" = "blue", "Bottom 10" = "red")) +  
+  scale_fill_manual(values = c("Top 10" = "steelblue", "Bottom 10" = "#ff5050")) +  
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
@@
 ```
-Figure 5 displays the average crime rate amongst the top 10 and bottom 10 priced neighbourhoods for each individual crime over the years 2014-2023.
+Figure \@ref(fig-crime-time-series-individuals) displays the average crime rates in the top 10 and bottom 10 priced neighborhoods for each individual crime from 2014 to 2023.
 
-- **Assault**: The assault rates remain consistently higher in Bottom 10 Lowest neighborhoods than in Top 10 Highest neighborhoods across all years.
-There seems to be a slight downward trend in the top neighborhoods around 2019–2021, while rates in lower neighborhoods remain relatively stable.
+**Upwards Trending Individual Crimes**
 
-- **Autotheft**: Bottom 10 Lowest neighborhoods show a steady increase in autotheft rates from 2014 onward, peaking around 2022.
-Top 10 Highest neighborhoods have significantly lower autotheft rates throughout the period, but they also show a slight upward trend, though the increase is far more gradual.
+- **Autotheft**: The top 10 neighborhoods show significantly higher auto theft rates from 2021 to 2023.
+Auto theft in the top 10 neighborhoods seems to follow an exponential trend, while the bottom 10 neighborhoods show a linear increase over time.
 
-- **Biketheft**: In the earlier years (2014–2017), Bottom 10 Lowest neighborhoods have higher bike theft rates, but Top 10 Highest neighborhoods catch up around 2018–2020.
-From 2020 onwards, the bike theft rates in Top 10 Highest neighborhoods are actually higher than those in the Bottom 10 Lowest neighborhoods.
+- **Assault**: Assault rates have consistently been higher in the bottom 10 neighborhoods compared to the top 10.
+However, the rate is growing fastest in the top 10 neighborhoods, nearly doubling from 2014 to 2023.
 
-- **Break and Enter**: Rates for break and enter crimes fluctuate more in Bottom 10 Lowest neighborhoods, peaking around 2016–2017 and then decreasing slightly.
-Top 10 Highest neighborhoods show more stable but much lower break and enter rates overall.
+- **Homicide**: Homicide rates remain low in both groups, but the bottom 10 neighborhoods exhibit slightly higher rates with a notable upward trend from 2016 to 2023.
+In contrast, homicide rates in the top 10 neighborhoods are more sporadic, with some years recording zero incidents.
 
-- **Homicide**: Both neighborhoods show very low homicide rates (as expected), but Bottom 10 Lowest neighborhoods exhibit slightly higher rates.
-There is a gradual upward trend in homicides in Bottom 10 Lowest neighborhoods from 2016 to 2023, although the scale is relatively small (0 to 7.5 per 100,000 residents).
+- **Theft over $5,000**: This crime is increasing in both neighborhood groups, with the top 10 neighborhoods consistently showing rates about 1.5 times higher than the bottom 10.
 
-- **Robbery**: Robbery rates in Bottom 10 Lowest neighborhoods are consistently higher than in Top 10 Highest neighborhoods.
-The gap between the two neighborhood types widens in 2020 and onwards, with a notable increase in robbery rates in Bottom 10 Lowest neighborhoods.
+- **Break and Enter**: Break and enter rates are significantly higher in the top 10 neighborhoods, reaching their highest levels in 2023, while the bottom 10 neighborhoods show a downward trend.
 
-- **Shooting**: Bottom 10 Lowest neighborhoods experience a significant peak in shooting rates around 2018, after which they decline.
-Top 10 Highest neighborhoods have almost negligible shooting rates in comparison.
+**Downwards Trending Individual Crimes**
 
-- **Theft from Motor Vehicle**: The theft from motor vehicle rates are much higher in Bottom 10 Lowest neighborhoods, and they fluctuate over the years, with no consistent trend.
-The rates for Top 10 Highest neighborhoods remain fairly stable and low across the entire time period.
+- **Robbery**: Robbery rates have consistently been higher in the bottom 10 neighborhoods, although the gap between the two groups has narrowed over time.
+The top 10 neighborhoods maintain a steady average of around 60, while the bottom 10 neighborhoods saw a decline from a peak of 200 in 2015 to a low of 100 in 2020, before starting to rise again, though still far below previous highs.
 
-- **Theft over $5,000**: This crime shows stable but relatively low rates for both neighborhood types, though Bottom 10 Lowest neighborhoods have a slight edge.
-There's an increasing trend in Bottom 10 Lowest neighborhoods starting in 2020.
+- **Shooting**: The bottom 10 neighborhoods experienced a significant peak in shooting rates around 2018, followed by a decline, though their rates remain consistently 8–10 times higher than those in the top 10 neighborhoods.
+
+- **Biketheft**: Bottom 10 neighbourhoods experience consistent bike theft rates, whereas top 10 neighbourhoods are significantly higher, pekaed in 2019 and have since signficantly decreased.
+
+- **Theft from Motor Vehicle**: Theft from motor vehicles shows similar rates in both neighborhood groups, but from 2020 to 2023, the top 10 neighborhoods have seen a significant decrease, while the bottom 10 neighborhoods have remained relatively high.
+
+Overall, the crimes most on the rise in the top 10 neighborhoods are auto theft, theft over $5,000, and break and enter.
+In the bottom 10 neighborhoods, homicide, auto theft, and assault are increasing the most.
+The crimes that have shown significant decreases in both groups are shootings and robbery.
+Theft from motor vehicles remains relatively stagnant, ranging from 200 to 300 incidents per 100,000 each year over the time period.
 
@@
 
-This analysis confirms the hypothesis that wealthier neighborhoods experience lower rates of violent crimes, while economically disadvantaged areas face greater exposure to violent crime.
-However, non-violent crimes seem more prevalent in wealthier neighborhoods, suggesting different socio-economic dynamics at play.
+This analysis confirms the hypothesis that wealthier neighborhoods experience lower rates of violent crimes, while economically disadvantaged areas face higher exposure to violent crime.
+However, non-violent crimes are more prevalent in wealthier neighborhoods, suggesting different socio-economic dynamics at play.
 
-The analysis of crime rates from 2014 to 2023 across Toronto's neighborhoods, focusing on the top 10 highest and bottom 10 lowest-priced neighborhoods, reveals significant disparities in crime patterns based on socioeconomic status.
-The results highlight a clear trend where violent crimes—such as assault, homicide, and shootings—are more prevalent in lower-priced neighborhoods, suggesting a direct correlation between economic disadvantage and exposure to violence.
-On the other hand, non-violent crimes like thefts (especially bike thefts and theft from motor vehicles) are more common in wealthier areas, reflecting different crime dynamics in these neighborhoods.
+The analysis of crime rates from 2014 to 2023 across Toronto's neighborhoods, focusing on the top 10 highest- and bottom 10 lowest-priced neighborhoods, reveals significant disparities based on socioeconomic status.
+Violent crimes—such as assault, homicide, and shootings—are more common in lower-priced neighborhoods, indicating a direct correlation between economic disadvantage and exposure to violence.
+Conversely, non-violent crimes like bike thefts and theft from motor vehicles are more frequent in wealthier areas, reflecting different crime dynamics.
 
@@
 
-- **Temporal Trends**: There is a city-wide spike in crime rates in 2023, following a period of lower crime during the 2020–2021 COVID-19 lockdowns, emphasizing the complex interactions between public health crises and crime patterns.
+- **Temporal Trends**: Crime rates city-wide spiked in 2023, following lower rates during the 2020–2021 COVID-19 lockdowns, highlighting the complex relationship between public health crises and crime patterns.
+Additionally, distinct trends are observed: auto theft, theft over $5,000, and break and enter surged in high-priced neighborhoods, while homicide, auto theft, and assault rose significantly in low-priced areas.
+Both groups saw decreases in shootings and robberies, while theft from motor vehicles remained relatively constant.
 
-These findings suggest that public policy should be tailored to address the specific needs of different economic groups.
-Low-income neighborhoods require increased focus on violence prevention, while higher-income areas might benefit from enhanced property crime deterrence.
-Moreover, the evidence points to the need for long-term, community-based crime reduction strategies that consider the broader social and economic factors contributing to crime.
+These findings suggest that public policy should be tailored to the specific needs of different economic groups.
+Low-income neighborhoods need a greater focus on violence prevention, while higher-income areas could benefit from enhanced property crime deterrence.
+Long-term, community-based crime reduction strategies that account for social and economic factors are essential.
+By incorporating temporal trends, policymakers can respond proactively to emerging crime patterns, rather than relying on static understandings of neighborhood crime dynamics.
 
-Future work should focus on expanding the scope of the analysis to include more socioeconomic indicators, such as employment rates and education levels, to better understand the root causes of crime and inform more effective policy interventions.
+Future research should expand the scope of analysis to include socioeconomic factors such as unemployment rates, education levels, and healthcare access, which likely influence crime rates.
+A broader understanding of the social and economic contexts in which crimes occur will allow for more effective, holistic interventions.
+Policies that integrate economic development, education, and mental health support, alongside safety measures, are key to achieving long-term reductions in both violent and non-violent crimes.
+
+In conclusion, this analysis emphasizes the need for differentiated crime prevention strategies that address the specific dynamics of each neighborhood.
+For low-income areas, tackling violence through social and economic interventions is crucial, while higher-income areas may require improved property crime deterrence.
+Public policy should aim to reduce socioeconomic disparities and ensure the safety and well-being of all residents.
+Temporal insights further underscore the importance of adapting policies to reflect changing crime patterns over time, ensuring that resources are allocated efficiently and interventions are responsive to evolving needs.
 
@@
 # Appendix {-}
+**Data Cleaning and Preparation**
+
+The data was cleaned using the janitor [@janitor], readxl [@readxl], tidyverse [@tidyverse] packages.
+
+- **Housing Price Data**: The housing data was prepared by filtering out invalid entries and ensuring that all price data was numeric.
+To investigate the potential relationship between socio-economic status and crime, the analysis focuses on the neighborhoods with the 10 highest and the 10 lowest average home prices.
+Thus, home price data for all other neighbourhoods were filtred out.
+These neighborhoods represent a wide socio-economic spectrum and are expected to exhibit varying patterns of criminal activity.
+By isolating these groups, we can assess whether wealthier neighborhoods experience lower crime rates and determine which types of crimes are more prevalent in less affluent areas.
+
+- **Crime Statistics Data**: The crime data was cleaned by removing any missing entries and focusing exclusively on the crime rates.
+Non-relevant columns (i.e., raw crime counts) were removed in favor of normalized rates per 100,000 residents, making comparisons across neighborhoods more meaningful.
 

### 2024-09-25 paper/paper.qmd
Added correct cross-referencing and added validation tests to the cleaned data script.

@@
 The analysis in this paper was conducted using R [@citeR].
-The paper is structured as follows: first, I review the selected data sources and their relevance to the subject.
-Second, I present key visualizations using the scales [@scales] and knitr [@knitr] packages to explore the relationships between neighborhood home prices, crime rates, and temporal trends from 2014 to 2023.
-Third, I provide an analysis of these visualizations and discuss their potential implications for public policy and future research into neighborhood-specific crime prevention strategies, based on the observed temporal trends in crime rates.
+The paper is structured as follows: first, I review the selected data sources and their relevance to the subject in @sec-data.
+Second, I present key visualizations using the scales [@scales] and knitr [@knitr] packages to explore the relationships between neighborhood home prices, crime rates, and temporal trends from 2014 to 2023 in @sec-results.
+Third, I provide an analysis of these visualizations in @sec-discussion and discuss their potential implications for public policy and future research into neighborhood-specific crime prevention strategies, based on the observed temporal trends in crime rates.
 
@@
 ### Summary Statistics of Housing Price Data 
-Table \@ref(tbl-housing-summary) an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods.
+Table @(tbl-housing-summary) an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods.
 
@@
 ```
+
+Table @(tbl-housing-summary) shows that the average home price in the top 10 high-priced neighborhoods is roughly five times higher than that of homes in the bottom 10 low-priced neighborhoods.
+Despite this large difference, the range of home prices in both neighborhood groups appears to have similar variability, with comparable standard deviations from their respective means.
+
 ### Mean Crime Rate Changes over Time
-Crime rate trends over the years (2014-2023) were analyzed for each crime type to understand patterns and shifts. This is visualized in Figure \@ref(fig-crime-comparison-0):
+The cleaned crime statistics dataset contains many more features than the home price dataset.
+For this reason, I present a visualization of the summary statistics in the form of a box plot for each individual crime in the dataset in the first year the data was collected (2014) vs the final year of data the dataset provides (2023).
+This is visualized in @(fig-crime-comparison-2014-2023):
 
@@
 #| fig-label: fig-crime-comparison-0
+#| out.width: "49%"
 #| show_col_types: FALSE
@@
   scale_y_continuous(labels = comma)  
+```
+
+```{r}
+#| fig-cap: "Crime Rate Distribution for 2014 and 2023 for Each Crime Type"
+#| fig-pos: "H"
+#| out.width: "49%"
+#| fig-label: fig-crime-comparison-2014-2023
+#| show_col_types: FALSE
+#| echo: false
+#| message: false
+
+# Load necessary packages
+library(tidyverse)
+
+# Read in the crime data
+crime_cleaned_data <- read_csv("../data/clean_data/crime_cleaned_data.csv")
+
+# Filter the data to include only 2014 and 2023
+crime_filtered <- crime_cleaned_data %>%
+  filter(year %in% c(2014, 2023))
+
+# Create box plots for each crime type for 2014 and 2023
+ggplot(crime_filtered, aes(x = factor(year), y = rate, fill = crime)) +
+  geom_boxplot() +
+  facet_wrap(~ crime, scales = "free_y") +  
+  labs(title = "Crime Rate Distribution for 2014 and 2023 for Each Crime Type",
+       x = "Year",
+       y = "Crime Rate per 100,000 Residents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
+  scale_y_continuous(labels = scales::comma)
 
 ```
-Figure \@ref(fig-crime-comparison-0) showcases a rather steady trend in crime rates with a dip in 2020 and 2021 which could be attributed to Covid-19 lockdowns, followed by a spike in 2023 of approximately 25% increase in mean crime rates across the city of Toronto's neighbourhoods.
+Figure @(fig-crime-comparison-0) highlights a relatively steady trend in crime rates, with a noticeable dip in 2020 and 2021, likely influenced by the COVID-19 lockdowns.
+This is followed by a sharp spike in 2023, marking an approximate 25% increase in the mean crime rates across Toronto’s neighborhoods.
+
+Figure @(fig-crime-comparison-2014-2023) provides detailed statistics on the rates of nine different crime types.
+Each crime category presents a significant number of outliers, which could potentially correspond to either the top 10 or bottom 10 neighborhoods—a hypothesis that will be further investigated.
+While some crime rates have significantly risen since 2014, others have remained stagnant or even declined.
+These trends will be thoroughly explored in the results section.
 
@@
 ```
-Figure \@ref(fig-crime-comparison-1) displays the mean crime rate for each crime type individually.
+Figure @(fig-crime-comparison-1) displays the mean crime rate for each crime type individually.
 There is a clear pattern of crime rates being at their lowest levels during 2020-2021 for all crime types, which can be argued to be due to Covid-19 lockdowns.
@@
 
-Figure \@ref(fig-crime-comparison) shows a comparative analysis of crime rates between the top 10 highest and bottom 10 lowest home price neighborhoods in Toronto from 2014 to 2023.
+Figure @(fig-crime-comparison) shows a comparative analysis of crime rates between the top 10 highest and bottom 10 lowest home price neighborhoods in Toronto from 2014 to 2023.
 
@@
 ```
-Figure \@ref(fig-crime-time-series) shows that the overall crime rate has increased at a much faster rate in the top 10 highest-priced neighborhoods compared to the bottom 10 lowest-priced neighborhoods.
+Figure @(fig-crime-time-series) shows that the overall crime rate has increased at a much faster rate in the top 10 highest-priced neighborhoods compared to the bottom 10 lowest-priced neighborhoods.
 The low-priced neighborhoods maintained relatively consistent crime rates around 1,500 per 100,000 residents until 2022, when the rate began to rise, reaching closer to 2,000.
@@
 ```
-Figure \@ref(fig-crime-time-series-individuals) displays the average crime rates in the top 10 and bottom 10 priced neighborhoods for each individual crime from 2014 to 2023.
+Figure @(fig-crime-time-series-individuals) displays the average crime rates in the top 10 and bottom 10 priced neighborhoods for each individual crime from 2014 to 2023.
 
@@
 
-# Discussion
+# Discussion {#sec-discussion}
 

### 2024-09-25 scripts/02-data_cleaning.R

@@ -1,41 +1,53 @@
 #### Preamble ####
-# Purpose: Cleans the raw housing data
+# Purpose: Cleans the raw housing and crime data and adds 'Group' classification for neighborhoods.
 # Author: Chenika Bukes
-# Date: 20 September 2024
+# Date: 24 September 2024
 # Contact: chenika.bukes@mail.utoronto.ca
 # License: MIT
 # Pre-requisites:
-#   - Installed the `tidyverse` and `janitor` R packages
-#   - Raw housing and crime data files saved as CSV or Excel
-# Any other information needed?
-# This script processes and cleans both housing price and crime rate data for further analysis.
+# - Installed `tidyverse` and `janitor` R packages
+# - Raw housing and crime data files saved as CSV or Excel
 
 #### Workspace setup ####
 library(tidyverse)
 library(janitor)
 library(readxl)
 
-#### Clean data ####
+#### Clean housing data ####
 # Read the third worksheet of the Excel file, which contains the housing data
 raw_data_housing <- read_excel("./data/raw_data/wellbeing-toronto-housing.xlsx", sheet = 3)
 
 # Cleaning steps:
-housing_cleaned_data <-
-  raw_data_housing |>
-  janitor::clean_names() |>
-  select(neighbourhood, neighbourhood_id, home_prices) |>
+housing_cleaned_data <- 
+  raw_data_housing %>%
+  janitor::clean_names() %>%
+  select(neighbourhood, neighbourhood_id, home_prices) %>%
   mutate(
     home_prices = as.numeric(gsub(",", "", home_prices)) # Remove commas and convert to numeric
-  ) |>
-  drop_na(home_prices) |>
-  filter(home_prices > 0) |>
+  ) %>%
+  drop_na(home_prices) %>%
+  filter(home_prices > 0) %>%
   rename(
     neighbourhood_name = neighbourhood,
     neighbourhood_code = neighbourhood_id,
     price = home_prices
   )
 
-#### Save cleaned housing data ####
+# Get the top 10 and bottom 10 neighborhoods based on home prices
+top_10 <- housing_cleaned_data %>%
+  arrange(desc(price)) %>%
+  slice(1:10) %>%
+  mutate(Group = "Top 10")
+
+bottom_10 <- housing_cleaned_data %>%
+  arrange(price) %>%
+  slice(1:10) %>%
+  mutate(Group = "Bottom 10")
+
+# Combine the top and bottom 10 neighborhoods into one dataset
[…]

### 2024-09-25 paper/paper.qmd
Paper finalised with updated cross references and appendix.

@@
 ### Summary Statistics of Housing Price Data 
-Table @(tbl-housing-summary) an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods.
+@tbl-housing-summary an overview of the housing price data for the top 10 highest and bottom 10 lowest-priced neighborhoods.
 
@@
 #| tbl-cap: "Summary Statistics of Housing Price Data (Top 10 Highest and Lowest)"
+#| label: tbl-housing-summary
 #| echo: false
@@
 
-Table @(tbl-housing-summary) shows that the average home price in the top 10 high-priced neighborhoods is roughly five times higher than that of homes in the bottom 10 low-priced neighborhoods.
+@tbl-housing-summary shows that the average home price in the top 10 high-priced neighborhoods is roughly five times higher than that of homes in the bottom 10 low-priced neighborhoods.
 Despite this large difference, the range of home prices in both neighborhood groups appears to have similar variability, with comparable standard deviations from their respective means.
@@
 #| fig-pos: "H"
-#| fig-label: fig-crime-comparison-0
+#| label: fig-crime-comparison-0
 #| out.width: "49%"
@@
 ```
-
-```{r}
-#| fig-cap: "Crime Rate Distribution for 2014 and 2023 for Each Crime Type"
-#| fig-pos: "H"
-#| out.width: "49%"
-#| fig-label: fig-crime-comparison-2014-2023
-#| show_col_types: FALSE
-#| echo: false
-#| message: false
-
-# Load necessary packages
-library(tidyverse)
-
-# Read in the crime data
-crime_cleaned_data <- read_csv("../data/clean_data/crime_cleaned_data.csv")
-
-# Filter the data to include only 2014 and 2023
-crime_filtered <- crime_cleaned_data %>%
-  filter(year %in% c(2014, 2023))
-
-# Create box plots for each crime type for 2014 and 2023
-ggplot(crime_filtered, aes(x = factor(year), y = rate, fill = crime)) +
-  geom_boxplot() +
-  facet_wrap(~ crime, scales = "free_y") +  
-  labs(title = "Crime Rate Distribution for 2014 and 2023 for Each Crime Type",
-       x = "Year",
-       y = "Crime Rate per 100,000 Residents") +
-  theme_minimal() +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
-  scale_y_continuous(labels = scales::comma)
-
-```
-Figure @(fig-crime-comparison-0) highlights a relatively steady trend in crime rates, with a noticeable dip in 2020 and 2021, likely influenced by the COVID-19 lockdowns.
+@fig-crime-comparison-0 highlights a relatively steady trend in crime rates, with a noticeable dip in 2020 and 2021, likely influenced by the COVID-19 lockdowns.
 This is followed by a sharp spike in 2023, marking an approximate 25% increase in the mean crime rates across Toronto’s neighborhoods.
-
-Figure @(fig-crime-comparison-2014-2023) provides detailed statistics on the rates of nine different crime types.
-Each crime category presents a significant number of outliers, which could potentially correspond to either the top 10 or bottom 10 neighborhoods—a hypothesis that will be further investigated.
-While some crime rates have significantly risen since 2014, others have remained stagnant or even declined.
-These trends will be thoroughly explored in the results section.
 
@@
 #| error: false
-#| fig-label: fig-crime-comparison-1
+#| label: fig-crime-comparison-1
 #| fig-cap: "Average Individual Crime Rates over all Neighbourhoods (2014-2023)"
@@
 ```
-Figure @(fig-crime-comparison-1) displays the mean crime rate for each crime type individually.
+
+@fig-crime-comparison-1 displays the mean crime rate for each crime type individually.
 There is a clear pattern of crime rates being at their lowest levels during 2020-2021 for all crime types, which can be argued to be due to Covid-19 lockdowns.
@@
 #| error: false
-#| fig-label: fig-crime-comparison
+#| label: fig-crime-comparison
 #| fig-cap: "Box Plots comparing Crime Rates in Top 10 vs Bottom 10 Home Price Neighborhoods"
@@
 
-Figure @(fig-crime-comparison) shows a comparative analysis of crime rates between the top 10 highest and bottom 10 lowest home price neighborhoods in Toronto from 2014 to 2023.
+@fig-crime-comparison shows a comparative analysis of crime rates between the top 10 highest and bottom 10 lowest home price neighborhoods in Toronto from 2014 to 2023.
 
@@
 #| error: false
-#| fig-label: fig-crime-time-series
+#| label: fig-crime-time-series
 #| fig-cap: "Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
@@
 ```
-Figure @(fig-crime-time-series) shows that the overall crime rate has increased at a much faster rate in the top 10 highest-priced neighborhoods compared to the bottom 10 lowest-priced neighborhoods.
+@fig-crime-time-series shows that the overall crime rate has increased at a much faster rate in the top 10 highest-priced neighborhoods compared to the bottom 10 lowest-priced neighborhoods.
 The low-priced neighborhoods maintained relatively consistent crime rates around 1,500 per 100,000 residents until 2022, when the rate began to rise, reaching closer to 2,000.
@@
 #| fig-pos: "H"
-#| fig-label: fig-crime-time-series-individuals
+#| label: fig-crime-time-series-individuals
 #| fig-cap: "Crime Rate Time Series from 2014 to 2023 for Top 10 vs Bottom 10 Housing Price Neighbourhoods"
@@
 ```
-Figure @(fig-crime-time-series-individuals) displays the average crime rates in the top 10 and bottom 10 priced neighborhoods for each individual crime from 2014 to 2023.
+@fig-crime-time-series-individuals displays the average crime rates in the top 10 and bottom 10 priced neighborhoods for each individual crime from 2014 to 2023.
 
@@
 Long-term, community-based crime reduction strategies that account for social and economic factors are essential.
-By incorporating temporal trends, policymakers can respond proactively to emerging crime patterns, rather than relying on static understandings of neighborhood crime dynamics.
+Temporal insights further underscore the importance of adapting policies to reflect changing crime patterns over time, ensuring that resources are allocated efficiently and interventions are responsive to evolving needs.
 
@@
 Policies that integrate economic development, education, and mental health support, alongside safety measures, are key to achieving long-term reductions in both violent and non-violent crimes.
-
-In conclusion, this analysis emphasizes the need for differentiated crime prevention strategies that address the specific dynamics of each neighborhood.
-For low-income areas, tackling violence through social and economic interventions is crucial, while higher-income areas may require improved property crime deterrence.
-Public policy should aim to reduce socioeconomic disparities and ensure the safety and well-being of all residents.
-Temporal insights further underscore the importance of adapting policies to reflect changing crime patterns over time, ensuring that resources are allocated efficiently and interventions are responsive to evolving needs.
 
@@
 
+**Crime Statistics Summary Visualization **
+
+@fig-crime-comparison-2014-2023 provides detailed statistics on the rates of nine different crime types.
+Each crime category presents a significant number of outliers, which could potentially correspond to either the top 10 or bottom 10 neighborhoods—a hypothesis that was further investigated and found to be true by the clearer interquartile ranges in @fig-crime-comparison.
+While some crime rates have significantly risen since 2014, others have remained stagnant or even declined.
+These trends were thoroughly explored in the results section.
+The outliers made it difficult to fully capture the patterns in the data and it rather inspired the necessary visualizations of the crime statistics, thus this was not kept in the final @sec-data.
+
+```{r}
+#| fig-cap: "Crime Rate Distribution for 2014 and 2023 for Each Crime Type"
+#| fig-pos: "H"
+#| label: fig-crime-comparison-2014-2023
+#| show_col_types: FALSE
+#| echo: false
+#| message: false
+#| warning: false
+#| error: false
+
+# Load necessary packages
+library(tidyverse)
+
+# Read in the crime data
+crime_cleaned_data <- read_csv("../data/clean_data/crime_cleaned_data.csv")
+
+# Filter the data to include only 2014 and 2023
+crime_filtered <- crime_cleaned_data %>%
+  filter(year %in% c(2014, 2023))
+
+# Create box plots for each crime type for 2014 and 2023
+ggplot(crime_filtered, aes(x = factor(year), y = rate, fill = crime)) +
+  geom_boxplot() +
+  facet_wrap(~ crime, scales = "free_y") +  
+  labs(title = "Crime Rate Distribution for 2014 and 2023 for Each Crime Type",
+       x = "Year",
+       y = "Crime Rate per 100,000 Residents") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  
+  scale_y_continuous(labels = scales::comma) +
+  theme(
+    plot.title = element_text(size = 10) 
+  )
+
+```
+
 \newpage
